### 2024-10-01 content/3u4_methods.tex
initial state from overleave

\section{Foundations of Ptolemaic Indexing}

\todo{explain Hetland's lower (upper) bound notation ($d^-$ and $d^+$)}
\begin{theorem}[Ptolemaic Bounds]
\label{thm4.2}
\todo{we might want to explain what the points are supposed to do before showing this theorem.
Or, alternatively, we just present this as a purely mathematical statement with no preceding motivation.}
    Let \(a, b, c, c' \in \mathbb{R}^n\) and let \(d(\cdot, \cdot)\) be the Euclidean distance. The distance \(d(a, c')\) can be bounded as follows:
    \begin{gather}
        \frac{1}{d(b, c)} \cdot \max \left\{
        \begin{array}{l}
            d(a, b)^- \cdot d(c, c') - d(a, c)^+ \cdot d(b, c'), \\
            d(a, c)^- \cdot d(b, c') - d(a, b)^+ \cdot d(c, c')
        \end{array}
        \right\} \leq d(a, c')
    \end{gather}
    and
    \begin{gather}
        d(a, c') \leq \frac{1}{d(b, c)} \cdot \left( d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c') \right)
    \end{gather}
\end{theorem}
\begin{proof}
    \todo{convert to passive voice}
    \todo{expand context to summarize all steps that lead to this proof}
    The lower bound corresponds to the derivation in equation \ref{eq4.4} and corresponds to the arrangement of the two pivot objects. Replace some distances with the respective lower or upper bound for the distance. The lower bounds can be used as they are always smaller than the exact distance. The upper bound must be used due to the subtraction. This means that the inequality still applies. The upper bound is obtained by using Ptolemy's inequality:

    \begin{equation*}
\begin{aligned}
\label{eq4.7}
    d(a,c') \cdot d(b,c) &\leq d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c') \\
    d(a,c') &\leq \frac{d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')}{d(b,c)}
\end{aligned}
\end{equation*}
    By replacing some distances with upper boundaries you get the desired upper boundaries.
\end{proof}

\section{Applying Ptolemy's Inequality to the Elkan's Algorithm}

\subsection{Integration in k-Means}
Theorem \ref{thm4.2} will now be applied in a similar context to Elkan's algorithm \ref{alg:elkan}, with the goal of integrating these new bounds into steps 5 and 6 of the algorithm.

Consider the data set \(D = \{x_1, \ldots, x_n\}\). For each \(x_i\), there is an upper \(u_i\) and lower bounds \(l_{ij}\) as in Elkan's algorithm. The new cluster center assigned in a iteration is denoted by \(c_i^{\text{new}}\), and the old cluster center by \(c_i^{\text{old}}\). To use Ptolemy's Inequality, one additional point is needed. The idea is to use an even older cluster center \(c_i^{\text{old,old}}\) obtained two iterations ago.

Overall, to obtain a new upper bound for the distance between the datapoint \(x_i\) and the new cluster center \(c_j^{\text{new}}\), consider the old cluster center \(c_j^{\text{old}}\) from the last iteration to which \(x_i\) was assigned, and the even older cluster center \(c_j^{\text{old,old}}\) from two iterations ago. Utilize the old upper bounds that were valid for the respective iteration.

\begin{equation*}
\begin{aligned}
    d(x_i,c_i^{old}) &\leq u_i^{old}\\
    d(x_i,c_i^{old, old}) &\leq u_i^{old, old}
\end{aligned}
\end{equation*}

Similarly, to obtain a new lower bound for the distance between the datapoint \(x_i\) and the new cluster center \(c_j^{\text{new}}\), the old lower bounds from the respective iterations are used:

\begin{equation*}
\begin{aligned}
    l_{i,j}^{\text{old}} &\leq d(x_i, c_j^{\text{old}}) \\
    l_{i,j}^{\text{old,old}} &\leq d(x_i, c_j^{\text{old,old}})
\end{aligned}
\end{equation*}

Using Theorem \ref{thm4.2}, the following upper bound for the distance are obtained:
\begin{equation}
\label{eq4.7}
    d(x_i, c_j^{\text{new}}) \leq u_i^{\text{new}} = \frac{1}{d(c_j^{\text{old}}, c_j^{\text{old,old}})} \cdot \left( u_i^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}) + u_i^{\text{old,old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old}}) \right)
\end{equation}

and the new lower bound:
\begin{equation}
\label{eq4.8}
    d(x_i, c_j^{\text{new}}) \geq l_{i,j}^{\text{new}} = \frac{1}{d(c_j^{\text{old}}, c_j^{\text{old,old}})} \cdot \max \left\{
        \begin{array}{l}
            l_{i,j}^{\text{old,old}} \cdot d(c_j^{\text{old}}, c_j^{\text{new}}) - u_i^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}), \\
            l_{i,j}^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}) - u_i^{\text{old,old}} \cdot d(c_j^{\text{old}}, c_j^{\text{new}})
        \end{array}
    \right\}
\end{equation}

You can see that the old upper boundaries are required to calculate the new lower boundaries.

Let us now look at how these bounds can be integrated into Elkan's algorithm. It is important to save not only the cluster centres and the bounds of the last iteration, but also those of the penultimate iteration.

At the beginning, all distances to all cluster centres are calculated in the first iteration. The upper bound is set by the distance to the nearest cluster centre, and the lower bounds are determined by the direct distances from the point to the respective cluster centres. An iteration is then performed as with Elkan, which means that bounds from two iterations are now available. This makes it possible to apply the new bounds from equations \ref{eq4.7} and \ref{eq4.8}. The algorithm then iterates, as in the original Elkan algorithm using updating with new update conditions, until the cluster centres converge or a specified number of iterations is reached. It is always important to check whether $d(c_j^{\text{old}}, c_j^{\text{old,old}}) \neq 0$. If this is the case, the update is carried out as for Elkan. The algorithm is displayed in Algorithm \ref{alg:ptolemy}.

\begin{algorithm}
\caption{Elkan's Algorithm extended with Ptolemy's inequality}
\label{alg:ptolemy}

\begin{algorithmic}

\State \textbf{Initialization:} Initialize all cluster centers. For each point $x_i$ and each center $c_j$, set the lower bound $l(x_i,c_j)$ and the upper bound $u(x_i)$. Assign each $x_i$ to the nearest cluster $C_j$ such that $c(x_i) = \arg \min_j d(x_i, c_j)$, utilizing Lemma 1 to minimize distance calculations. Set $r(x_i) = \text{true}$ for all points.

\State \textbf{Step 0:} Compute one iteration of Elkan's algorithm. Save old upper and lower bounds of the last two iterations.
\Repeat
    \State \textbf{Step 1:}  Compute distances $d(c_i, c_j)$ between all centers, and calculate $s(c_i) = \frac{1}{2} \min_{c_j \neq c_i} d(c_i, c_j)$ for each center $c_i$.
    \Statex
    \State \textbf{Step 2:}  Retain points $x_i$ in their current clusters if $u(x_i) \leq s(c(x_i))$.
    \Statex
    \State \textbf{Step 3:}  For remaining points, consider $x_i$ for reassignment if:
        \begin{itemize}
            \item $c_j \neq c(x_i)$,
            \item $u(x_i) > l(x_i, c_j)$, and
            \item $u(x_i) > \frac{1}{2}d(c(x_i), c_j)$.
        \end{itemize}
    \Statex
    \State \textbf{Step 3a:}  If $r(x_i)$ is true, compute $d(x_i, c(x_i))$. Set $r(x_i) = \text{false}$. Otherwise, $u(x_i) =d(x_i, c(x_i))$
    \Statex
    \State \textbf{Step 3b:} If $d(x_i, c(x_i)) > l(x_i, c_j)$ or $d(x_i, c(x_i)) > \frac{1}{2}d(c(x_i), c_j)$, compute $d(x_i, c_j)$. Reassign $x_i$ to $C_j$ if $d(x_i, c_j) < d(x_i, c(x_i))$.
    \Statex
    \State \textbf{Step 4:} Compute the cluster centers as the centroids of the corresponding clusters $c'_j$.
    \Statex
    \State \textbf{Step 5:} If distance between two older cluster is not 0. Update lower bounds by using equation \ref{eq4.8}. Otherwise, update like in Elkan. Save the old lower bounds of the two last iterations.
    \Statex
    \State \textbf{Step 6:} If distance between two older cluster is not 0. Update upper bounds by using equation \ref{eq4.7}. Otherwise, update like in Elkan. Reset $r(x_i) = \text{true}$. Save old cluster and upper bounds of the last two iterations.
    \Statex
    \State \textbf{Step 7:} Replace each center $c_j$ with $c'_j$.

\Until{convergence}

\end{algorithmic}
\end{algorithm}

### 2024-10-01 main.tex
initial state from overleave

% !TeX encoding = UTF-8
% !TeX program = pdflatex
% !BIB program = bibtex

\documentclass[english]{lni}

\input{content/preamble}

\begin{document}
%%% Mehrere Autoren werden durch \and voneinander getrennt.
%%% Die Fußnote enthält die Adresse sowie eine E-Mail-Adresse.
%%% Das optionale Argument (sofern angegeben) wird für die Kopfzeile verwendet.
\title[Ein Kurztitel]{Extending K-Means Clustering with Ptolemy’s Inequality}
%% \subtitle{Untertitel / Subtitle} % if needed
 \author[1,2]{Max Pernklau}{firstname1.lastname1@affiliation1.org}{0000-0000-0000-0000}
 \author[2]{Nikita Averitchev}{firstname2.lastname2@affiliation2.org}{0000-0000-0000-0000}
 \author[3]{Christian Beecks}{firstname3.lastname3@affiliation1.org}{0000-0000-0000-0000}
% \author[1]{Firstname4 Lastname4}{firstname4.lastname4@affiliation1.org}{0000-0000-0000-0000}%
 \affil[1]{Universität 1\\Abteilung\\Straße\\Postleitzahl Ort\\Land}
 \affil[2]{University 2 \\Department\\Address\\Country}
 \affil[3]{University 3\\Department\\Address\\Country}
\maketitle

\begin{abstract}
Clustering is a fundamental data-analytical operation in the field of unsupervised learning. Given a database of unknown structure, clustering aims to discover the inherent structure of the data objects according to similarity, such that similar data objects are grouped together and dissimilar ones are separated in different groups or clusters. %The resulting groups, which are denoted as clusters, then represent the clustering structure of the database. A
Among the various approaches of clustering methods, we focus on partitioning methods, i.e. on the specific problem of \emph{k-means clustering}, which minimizes the intra-cluster variance. 
%where the Lloyd algorithm is a prominent and de facto standard implementation.
As the standard algorithmic approach to k-means clustering, namely the Lloyd algorithm, lacks on efficiency and scalability, various adaptations and modifications have been developed yielding to the family of fast k-means clustering algorithms. In this short paper, we extend the clustering algorithm proposed by Elkan with Ptolemy's inequality, which generalizes the triangle inequality, to prune unnecessary distance calculations. The aim of our short paper is not to compete with the state-of-the-art algorithmic solutions for the k-means clustering problem, but to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations.

\end{abstract}
\begin{keywords}
k-means clustering \and Lloyd algorithm \and Ptolemy's inequality %Keyword1 \and Keyword2
\end{keywords}
%%% Beginn des Artikeltexts

\todo{Americanize spelling (thesis uses chiefly British terms)}

\section{Introduction}

%intro
Clustering is a fundamental data-analytical operation in the domain of data science, where the basic task is to divide a set of data objects into different groups or clusters, such that objects within the same cluster have sufficient similarities, while objects in different clusters have significant differences. The collection of clusters reflects the inherent structure of the database and is denoted as clustering.

Clustering can be found in various scientific or economic application domains \cite{ezugwu2022comprehensive,oyewole2023data, gan2020data}, ranging from archaeology \cite{troiano2024comparative} and finance \cite{cai2016clustering} to industry \cite{lee2021technological} and zoology \cite{shen2021multivariate}, to name just a few. Especially in the field of unsupervised learning, clustering is utilized to extract structures from unlabeled data \cite{chander2023data}.

Alongside the diverse applications of clustering, unique domain-specific requirements and challenges arise, each of which can be addressed by leveraging different families of clustering approaches \cite{xu2015comprehensive,han2012data}. These approaches, ranging from partitioning and hierarchical methods to density-based, grid-based and graph-based methods, are designed to accommodate varying data characteristics and application-related requirements to facilitate efficient cluster analyses across diverse domains.

%different algorithms/problems
%kmeans/lloyd
%fast algorithms
%inefficient: distance calculations
%our proposal: ptolemaic
%our purpose

In this short paper, we focus on partitioning methods due to their simplicity in terms of interpretability and implementability \cite{DBLP:conf/iiwas/BeecksBHLSD22}. In this field, the \emph{k-means algorithm} \cite{bock2007clustering,hans2008origins,DBLP:journals/prl/Jain10,steinley2006k} has become one of the most influential clustering techniques \cite{DBLP:journals/kais/WuKQGYMMNLYZSHS08,olukanmi2019rethinking}. Although the term "k-means algorithm" is explicitly credited to MacQueen \cite{macqueen1967}, its origins trace back to Steinhaus \cite{steinhaus1956division} in 1956, with its first application to data clustering by Forgy \cite{forgy1965cluster} in 1965. The widely recognized version in use today is the \emph{Lloyd algorithm} \cite{DBLP:journals/tit/Lloyd82}, introduced in 1982.

With the proliferation of complex data sources and the growing demand for more efficient clustering techniques, numerous \emph{fast k-means algorithms} have been introduced \cite{DBLP:conf/icml/Elkan03,DBLP:conf/sdm/Hamerly10,drake2012accelerated,hamerly2015accelerating,DBLP:conf/icml/NewlingF16,DBLP:conf/icml/DingZSMM15}. These methods are designed to achieve exact clustering results with significantly reduced computational cost. Moreover, they do not require any form of precomputation, making them directly applicable across a wide range of scenarios.

%contribution

%structure

\newpage

\section{Related Work}

\input{content/3u4_methods}

\section{Preliminary Results}

\begin{figure}
  \centering
  \includegraphics[width=.8\textwidth]{fig/gaussia+random.png}
  \caption{}
  \label{fig:gauss_univariate}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.8\textwidth]{fig/wine+iris.png}
  \caption{}
  \label{fig:data}
\end{figure}

\todo{
    Experiments to show:
    - iris, wine, gaussian, uniform
    (moons, circles aren't a good fit for k-means -- is that reason enough to exclude them?)
}

\todo{we should mention that the Pto\_low version performs worse in every test (but one), but we do not need to include it in the comparison between Elkan and Our Work}

\todo{compile a graph that summarizes the four pages of tables}

\todo{How much do we "trust" these results, provided that each test configuration ( [k choice]x[dataset]x[algorithm]) was only run once?
We might need to repeat and average over these experiments.
TBD
}

\section{Conclusions and Future Work}

%%% Angabe der .bib-Datei (ohne Endung) / State .bib file (im Falle der Nutzung von BibTeX)
\bibliography{references,bibliography_thesis,bibliography_iiwas}
%% \printbibliography % im Falle der Nutzung von biblatex
\end{document}

### 2024-10-01 content/3u4_methods.tex
format

@@
 			\label{eq4.7}
-    d(a,c') \cdot d(b,c) &\leq d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c') \\
-    d(a,c') &\leq \frac{d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')}{d(b,c)}
+			d(a,c') \cdot d(b,c) & \leq d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')                \\
+			d(a,c')              & \leq \frac{d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')}{d(b,c)}
 		\end{aligned}
@@
 	\begin{aligned}
-    d(x_i,c_i^{old}) &\leq u_i^{old}\\
-    d(x_i,c_i^{old, old}) &\leq u_i^{old, old}
+		d(x_i,c_i^{old})      & \leq u_i^{old}      \\
+		d(x_i,c_i^{old, old}) & \leq u_i^{old, old}
 	\end{aligned}
@@
 	\begin{aligned}
-    l_{i,j}^{\text{old}} &\leq d(x_i, c_j^{\text{old}}) \\
-    l_{i,j}^{\text{old,old}} &\leq d(x_i, c_j^{\text{old,old}})
+		l_{i,j}^{\text{old}}     & \leq d(x_i, c_j^{\text{old}})     \\
+		l_{i,j}^{\text{old,old}} & \leq d(x_i, c_j^{\text{old,old}})
 	\end{aligned}

### 2024-10-01 content/3u4_methods.tex
compressing clustering, k-means definitions
progress just before talk with Krumpel

@@
 \section{Foundations of Ptolemaic Indexing}
+
+To discuss the k-means problem, we first need to formalize the concept of a \emph{clustering algorithm}:
+\begin{definition}[clustering algorithm]
+	Given a target number of clusters $k$,
+	let $F$ be a function that assigns each element from an input data set $D$
+	cluster index $i \in \{1, 2, \ldots, k\}$
+	$$ F:D \to \{1, 2, \ldots, k\} \;.$$
+\end{definition}
+Notably, we restrict our analysis to clustering problems that use the $n$-dimensional Euclidean space as the domain of the input data set: $D \subset \mathbb{R}^n$.
+This also implies that the similarity of the data points to be clustered is completely defined by the Euclidean distance $d(x,y)= || x-y ||_2 $.
+
+The heart of the k-means problem is the assignment of clusters so that the similarity variance between all points in the same cluster should be reasonably small.
+More formally:
+\begin{definition}[$k$-means problem]
+	Let $x_i \in D$ be an element from the input data set of a clustering algorithm $F$
+	and let $C_i$ be the set of all elements assigned to cluster $i$:
+	$$ C_i = \{x \in D \mid F(x) = i\} \,.$$
+	Furthermore, let $E$ be the sum of inter-cluster variances
+	\begin{align}
+		E & = \sum_{i=1}^k \operatorname{Var}[C_i] = \sum_{i=1}^k \sum_{x_j \in C_i}  d(x_j, c_i)\,,
+	\end{align}
+	where $c_i = \operatorname{E}[C_i]$ is the mean over all points in the cluster.
+
+	Then, the map $F$ is called a solution to the $k$-means problem
+	if $E$ cannot be made smaller by changing the assignment of any one element to a different cluster.
+\end{definition}
+\todo{Note that this is a local minimum definition that the Lloyd algorithm can find}
 

### 2024-10-02 content/3u4_methods.tex
proofreading and prelims up to Elkan

--- 2024-10-01 content/3u4_methods.tex
@@
-\section{Foundations of Ptolemaic Indexing}
+\section{Preliminaries}
+\newcommand{\kmeans}{$k$-means problem\xspace}
+
+\note{explain structure shortly}
+
+\subsection{The \kmeans}
+
+To discuss the \kmeans, we first need to formalize the concept of a \emph{clustering algorithm}:
+\begin{definition}[clustering algorithm]
+	Given a target number of clusters $k$,
+	let $F$ be a function that assigns each element from an input data set $D$ to a
+	cluster index $i \in \{1, 2, \ldots, k\}$
+	$$ F:D \to \{1, 2, \ldots, k\} \;.$$
+\end{definition}
+Notably, we restrict our analysis to clustering problems that use the $n$-dimensional Euclidean space as the domain of the input data set: $D \subset \mathbb{R}^n$.
+Similarly, we assume that the similarity of the data points to be clustered is completely defined by the Euclidean distance $d(x,y)= || x-y ||_2 $.
+The rationale behind this assumption will be made clear shortly in \autoref{pto}.
+
+The heart of the \kmeans is the assignment of clusters so that the variance in similarity between all points in the same cluster should be reasonably small.
+More formally:
+\begin{definition}[\kmeans]
+	Let $F$ be a clustering algorithm
+	%$x \in D$ be an element from the input data set of a clustering algorithm $F$
+	and let $C_i$ be the set of all elements assigned to cluster $i$:
+	$$ C_i = \{x \in D \mid F(x) = i\} \,.$$
+	Furthermore, let $\epsilon$ be the weighted sum of inter-cluster variances
+	$$ \epsilon = \sum_{i=1}^k |C_i| \operatorname{Var}[C_i] = \sum_{i=1}^k \sum_{x_j \in C_i}  d^2(x_j, c_i)\,, $$
+	where $c_i = \operatorname{E}[C_i]$ is the mean over all points in the cluster.
+
+	Then, the map $F$ is called a solution to the \kmeans
+	if $\epsilon$ cannot be made smaller by changing the assignment of any one element to a different cluster.
+\end{definition}
+Note that with this definition, the common $k$-means algorithms, like Lloyd's, provides solution to the \kmeans,
+as we do not require that $F$ finds the smallest possible $\epsilon$,
+just a local minimum of $\epsilon$.
+This is consistent with the colloquial notion that solutions to the \kmeans ``can be found'',
+even when an optimal solution (an NP-hard problem) is not practically obtainable \cite{}.
+
+The most prevalent algorithm employed to identify valid solutions to the \kmeans is the Lloyd algorithm.
+For convenience, it is reproduced in \autoref{alg:lloyd};
+given its widespread use, we will only cover its most significant characteristics as they relate to our analysis.
+
+\begin{algorithm}
+	\caption{k-Means Algorithm}
+	\label{alg:lloyd}
+
+	\textbf{Input:} \( k \): Number of clusters, \( D \): Dataset containing \( n \) objects
+
+	\textbf{Output:} Assignment of each $x_i \in D$ to a clusters
+
+	\begin{algorithmic}[1]
+		\State Initialize \( k \) cluster centers \( \{c_1, c_2, \dots, c_k\} \) arbitrarily from \( D \)
+		\Repeat
+		\label{algstep:assign}
+		\State \textcolor{gray}{// Assign \( x_i \) to the closest cluster}
+		\For{each object \( x_i \) in \( D \)}
+		\State \( C_j \leftarrow C_j \cup \{x_i\} \) where \( j = \underset{j}{\arg\min} \; d^2(x_i, c_j) \)
+		\EndFor
+		\State \textcolor{gray}{// Update cluster center \( c_j \)}
+		\For{each cluster \( C_j \)}
+		\State \( c_j \leftarrow \frac{1}{|C_j|} \sum_{x_i \in C_j} x_i \)
+		\EndFor
+		\Until{no change in cluster assignments}
+	\end{algorithmic}
+\end{algorithm}
+
+For each iteration of Lloyds algorithm, the assignment step (\autoref{algstep:assign}) needs to identify the closest cluster center for each element from the input dataset $D$.
+To do so, Lloyd's algorithm calculates the distances between each cluster and each element from $D$, requiring $k\cdot|D|$ total distance evaluations per iteration.
+These frequent distance evaluations can make up a significant share of the algorithms computational costs, especially for high-dimensional datasets.
+
+\subsection{Accelerating $k$-means}
+
+As we have seen, Lloyd's algorithm requires numerous evaluations of the distance function.
+However, the exact distances do not actually need to be calculated explicitly for every element, as it is sufficient to identify which cluster center is the closest to a given element.
+This is often possible from geometric considerations alone, when lower and upper distance bounds are available.
+Two properties of the Euclidean space are of particular use here:
+The triangle inequality
+$$ d(x,y) \leq d(x,z) + d(z,y) $$
+and Ptolemy's inequality
+$$ d(x, y)\cdot d(v, u) \leq d(x, v) \cdot d(y,v) + d(x, u) \cdot d(y, v)$$
+can provide lower and upper bound on the distance between two points given two (respectively five) other distances.
+
+Numerous algorithms have been developed that exploit the former inequality to avoid explicit distance evaluations.
+However, to the best of our knowledge, the latter has yet to be employed for the purposes of accelerating k-means clustering.
+
+Therefore, we define Elkan's algorithm \cite{}, which only uses the triangle inequality, here and extend it by incorporating Ptolemy's inequality in \ref{}.
+
+\todo{insert Elkan's algo here? Or can we do it in any other way to save some space?}
 

### 2024-10-02 content/5_results.tex
changes from overleaf, mostly NiAv

The developed variant of the k-Means algorithm was tested by implementing the classical k-Means algorithm (Lloyd's algorithm), Elkan's algorithm, and the Ptolemy-based extension in Python. The code is available on GitHub~\cite{Averitchev2024kMeansPtolemy}. To provide a fair comparison, a variable was included to count the number of distance calculations for each algorithm. This metric was chosen because distance calculations account for the majority of computing time in k-Means algorithms, offering a clear measure of efficiency regardless of implementation optimizations.

Additionally, two intermediate variants were implemented to assess the separate effects of the lower and upper bounds: one uses Elkan's lower bound with Ptolemy's upper bound, and the other uses Ptolemy's lower bound with Elkan's upper bound.

\subsection{Datasets}

A variety of datasets were used to compare the performance of the algorithms. The datasets are summarized in Table~\ref{tab:datasets}. 
\begin{table}[htbp]
\centering
\caption{Datasets Used in the Experiments}
\label{tab:datasets}
\begin{tabular}{lll}
\hline
\textbf{Dataset} & \textbf{Cardinality} & \textbf{Dimensionality} \\
\hline
\multicolumn{3}{l}{\textit{Fixed-Structure Datasets}} \\
Iris & 150 & 4 \\
Wine & 178 & 13 \\
\hline
\multicolumn{3}{l}{\textit{Variable-Dimension Datasets}} \\
Gaussian Low Dimension & 10,000 & 10 \\
Gaussian Medium Dimension & 10,000 & 100 \\
Gaussian High Dimension & 10,000 & 300 \\
Random Low Dimension & 5,000 & 10 \\
Random Medium Dimension & 5,000 & 100 \\
Random High Dimension & 5,000 & 300 \\
\hline
\end{tabular}
\end{table}

The fixed-structure datasets include the Iris and Wine datasets \cite{pedregosa2011scikit}, which have fixed dimensions and cardinalities and represent different clustering structures. The variable-dimension datasets include Gaussian and Uniform Random datasets, which offer flexibility in terms of dimensionality and cardinality, making them ideal for testing the algorithms under various conditions.

\subsection{Experiments}

The experiments were designed to evaluate the algorithms based on the number of distance calculations, which dominate the computing time in k-Means algorithms. This metric ensures comparability across different implementations and hardware setups. Each algorithm was executed three times on each dataset, with the parameters $k = 3$, $k = 20$, and $k = 100$.

The results for the Iris and Wine datasets are visualized in Fig. \ref{fig:wine+iris}, while Fig. \ref{fig:gauss_univariate} illustrates the performance on the Random and Gaussian dataset variations. These figures present the speedup relative to Elkan's algorithm, showcasing any improvements introduced by the proposed extension.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig/combined_plot.pdf}
  \caption{}
  \label{fig:combined}
\end{figure}

\subsection{Analysis of the Results}

The experimental results, presented in the figures, provide critical insights into the performance of the k-Means algorithm variants. The primary focus of this analysis is on the full Ptolemaic algorithm and its comparison with Elkan’s algorithm. The results highlight several scenarios in which the full Ptolemaic algorithm achieves substantial improvements over Elkan’s approach, particularly in reducing the number of distance evaluations.

\textbf{Influence of the Number of Clusters}

The full Ptolemaic algorithm consistently demonstrates notable improvements over Elkan’s algorithm as the number of clusters ($k$) increases. For instance, in the Gaussian low-dimensional dataset, the Ptolemaic algorithm achieves speedups of over 2-fold for $k = 20$ and surpasses 3-fold for $k = 100$, illustrating its ability to significantly reduce distance calculations in more complex clustering tasks. Similar trends are observed in the Random low-dimensional dataset, where the full Ptolemaic algorithm achieves a speedup of more than 3-fold at $k = 20$ and $k = 100$. These results underscore the increasing efficiency of the Ptolemaic algorithm as the clustering complexity grows.

In contrast, for smaller cluster counts, such as $k = 3$, the performance advantage of the full Ptolemaic algorithm is less pronounced but still evident, with speedups of around 1.1 in both the Iris and Wine datasets. This suggests that while Ptolemy’s inequality offers significant improvements for larger $k$ values, it still provides moderate gains for smaller cluster sizes.

The Ptolemaic Upper Bound variant performs comparably to the full Ptolemaic algorithm, especially for larger $k$ values, delivering similar efficiency improvements. However, the Ptolemaic Lower Bound variant tends to underperform in some cases, particularly with smaller cluster counts or in higher-dimensional datasets, where its speedup is sometimes close to or even below that of Elkan’s algorithm.

\textbf{Influence of Dimensionality}

The full Ptolemaic algorithm also performs well across datasets with varying dimensionality. In the Gaussian low-dimensional dataset (10 dimensions), the full Ptolemaic algorithm achieves impressive speedups, particularly at higher cluster counts, with speedups over 3-fold at $k = 100$. As the dimensionality increases, the performance gains of the Ptolemaic algorithm persist, though they become slightly less pronounced. For example, in the Gaussian medium (100 dimensions) and Gaussian high (300 dimensions) datasets, the Ptolemaic algorithm continues to offer significant speedups, but the performance advantage over Elkan’s algorithm decreases slightly.

In higher-dimensional datasets, such as the Random high-dimensional dataset, the full Ptolemaic algorithm remains effective, consistently reducing the number of distance evaluations, particularly for larger $k$ values. However, the Ptolemaic Lower Bound variant often struggles in these higher-dimensional cases, further emphasizing the robustness of the full Ptolemaic approach.

\textbf{Summary of Results}

The full Ptolemaic algorithm demonstrates consistent and significant improvements over Elkan’s algorithm across a variety of datasets, particularly in scenarios with a high number of clusters and low to medium dimensionality. The Ptolemaic Upper Bound variant performs similarly in many cases, offering comparable efficiency gains, whereas the Ptolemaic Lower Bound variant sometimes struggles, particularly in datasets with higher dimensionality or smaller cluster counts. Overall, the full Ptolemaic algorithm proves to be a robust and efficient alternative to traditional k-Means implementations, especially as the complexity of the clustering task increases.

### 2024-10-02 content/5_results.tex
proofread first part of results

@@
-The developed variant of the k-Means algorithm was tested by implementing the classical k-Means algorithm (Lloyd's algorithm), Elkan's algorithm, and the Ptolemy-based extension in Python.
+\todo{If we claim to have implemented the classic k-means, we also have to show that in our plots.}
+Our variant of the k-Means algorithm is tested by implementing the classical k-Means algorithm (Lloyd's algorithm), Elkan's algorithm, and its novel, Ptolemy-based extension in Python.
 The code is available on GitHub~\cite{Averitchev2024kMeansPtolemy}.
-To provide a fair comparison, a variable was included to count the number of distance calculations for each algorithm.
-This metric was chosen because distance calculations account for the majority of computing time in k-Means algorithms, offering a clear measure of efficiency regardless of implementation optimizations.
+To provide a fair comparison of the algorithms,
+we measure the number of distance calculations instead of wall-clock time.
+This metric was chosen as distance calculations account for the majority of computing time in k-Means algorithms, offering a clear measure of efficiency regardless of implementational optimizations.
 
-Additionally, two intermediate variants were implemented to assess the separate effects of the lower and upper bounds: one uses Elkan's lower bound with Ptolemy's upper bound, and the other uses Ptolemy's lower bound with Elkan's upper bound.
+Additionally, two intermediate variants were implemented to assess the separate effects of the lower and upper bounds:
+\emph{Ptolemy\_upper} one uses Elkan's lower bound with Ptolemy's upper bound, and \emph{Ptolemy\_lower} uses Ptolemy's lower bound with Elkan's upper bound.
 
 \subsection{Datasets}
+We employ two groups of datasets to compare the different algorithms, summarized in Table~\ref{tab:datasets}:
+The two real-world datasets include the Iris and Wine datasets \cite{pedregosa2011scikit}, which have fixed dimensions and cardinalities and represent different clustering structures.
+The two synthetic datasets include Gaussian and uniformly randomly generated points, which are used to test for the effects of changing dataset dimensionality.
 
-A variety of datasets were used to compare the performance of the algorithms. The datasets are summarized in Table~\ref{tab:datasets}. 
-\begin{table}[htbp]
+\begin{table}[bp]
 	\centering
@@
 	\label{tab:datasets}
-\begin{tabular}{lll}
-\hline
-\textbf{Dataset} & \textbf{Cardinality} & \textbf{Dimensionality} \\
-\hline
-\multicolumn{3}{l}{\textit{Fixed-Structure Datasets}} \\
-Iris & 150 & 4 \\
-Wine & 178 & 13 \\
-\hline
-\multicolumn{3}{l}{\textit{Variable-Dimension Datasets}} \\
-Gaussian Low Dimension & 10,000 & 10 \\
-Gaussian Medium Dimension & 10,000 & 100 \\
-Gaussian High Dimension & 10,000 & 300 \\
-Random Low Dimension & 5,000 & 10 \\
-Random Medium Dimension & 5,000 & 100 \\
-Random High Dimension & 5,000 & 300 \\
-\hline
+	\begin{tabular}{llSS}
+		\toprule
+		\textbf{Group}                                    & \textbf{Dataset}          & \textbf{Cardinality} & \textbf{Dimensionality} \\
+		\midrule
+		\multirow{2}{*}{\textit{fixed structure}}         & Iris                      & 150                  & 4                       \\
+		                                                  & Wine                      & 178                  & 13                      \\
+		\cmidrule{2-4}
+		\multirow{6}{*}{\textit{variable dimensionality}} & Gaussian Low Dimension    & 10000                & 10                      \\
+		                                                  & Gaussian Medium Dimension & 10000                & 100                     \\
+		                                                  & Gaussian High Dimension   & 10000                & 300                     \\
+		                                                  & Random Low Dimension      & 5000                 & 10                      \\
+		                                                  & Random Medium Dimension   & 5000                 & 100                     \\
+		                                                  & Random High Dimension     & 5000                 & 300                     \\
+		\bottomrule
 	\end{tabular}
@@
 
-The fixed-structure datasets include the Iris and Wine datasets \cite{pedregosa2011scikit}, which have fixed dimensions and cardinalities and represent different clustering structures.
-The variable-dimension datasets include Gaussian and Uniform Random datasets, which offer flexibility in terms of dimensionality and cardinality, making them ideal for testing the algorithms under various conditions.
-
 \subsection{Experiments}
-
 The experiments were designed to evaluate the algorithms based on the number of distance calculations, which dominate the computing time in k-Means algorithms.

### 2024-10-02 main.tex
changes from overleaf, mostly NiAv
proofread first part of results
further proofing, problem with bibliography

@@
 % !BIB program = bibtex
-
-\documentclass[english]{lni}
 
@@
 In this field, the \emph{k-means algorithm} \cite{bock2007clustering,hans2008origins,DBLP:journals/prl/Jain10,steinley2006k} has become one of the most influential clustering techniques \cite{DBLP:journals/kais/WuKQGYMMNLYZSHS08,olukanmi2019rethinking}.
-Although the term "k-means algorithm" is explicitly credited to MacQueen \cite{macqueen1967}, its origins trace back to Steinhaus \cite{steinhaus1956division} in 1956, with its first application to data clustering by Forgy \cite{forgy1965cluster} in 1965.
+Although the term \emph{k-means algorithm} is explicitly credited to MacQueen \cite{macqueen1967}, its origins trace back to Steinhaus \cite{steinhaus1956division} in 1956, with its first application to data clustering by Forgy \cite{forgy1965cluster} in 1965.
 The widely recognized version in use today is the \emph{Lloyd algorithm} \cite{DBLP:journals/tit/Lloyd82}, introduced in 1982.
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\textwidth]{fig/gaussia+random.png}
-  \caption{}
-  \label{fig:gauss_univariate}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\textwidth]{fig/wine+iris.png}
-  \caption{}
-  \label{fig:data}
-\end{figure}
+\input{content/5_results}
 
@@
 
+\clearpage
 %%% Angabe der .bib-Datei (ohne Endung) / State .bib file (im Falle der Nutzung von BibTeX)
-\bibliography{references,bibliography_thesis,bibliography_iiwas}
+\todo{re-enable citations}
+%\bibliography{references,bibliography_thesis,bibliography_iiwas}
 %% \printbibliography % im Falle der Nutzung von biblatex

### 2024-10-02 content/alg_elkan2.tex
break out and typographically improve algo. Use eq refs in the critical steps

--- 2024-10-01 content/3u4_methods.tex
@@
-\section{Foundations of Ptolemaic Indexing}
-
-\todo{explain Hetland's lower (upper) bound notation ($d^-$ and $d^+$)}
-\begin{theorem}[Ptolemaic Bounds]
-	\label{thm4.2}
-	\todo{we might want to explain what the points are supposed to do before showing this theorem.
-		Or, alternatively, we just present this as a purely mathematical statement with no preceding motivation.}
-	Let \(a, b, c, c' \in \mathbb{R}^n\) and let \(d(\cdot, \cdot)\) be the Euclidean distance. The distance \(d(a, c')\) can be bounded as follows:
-	\begin{gather}
-		\frac{1}{d(b, c)} \cdot \max \left\{
-		\begin{array}{l}
-			d(a, b)^- \cdot d(c, c') - d(a, c)^+ \cdot d(b, c'), \\
-			d(a, c)^- \cdot d(b, c') - d(a, b)^+ \cdot d(c, c')
-		\end{array}
-		\right\} \leq d(a, c')
-	\end{gather}
-	and
-	\begin{gather}
-		d(a, c') \leq \frac{1}{d(b, c)} \cdot \left( d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c') \right)
-	\end{gather}
-\end{theorem}
-\begin{proof}
-	\todo{convert to passive voice}
-	\todo{expand context to summarize all steps that lead to this proof}
-	The lower bound corresponds to the derivation in equation \ref{eq4.4} and corresponds to the arrangement of the two pivot objects.
-Replace some distances with the respective lower or upper bound for the distance.
-The lower bounds can be used as they are always smaller than the exact distance.
-The upper bound must be used due to the subtraction.
-This means that the inequality still applies.
-The upper bound is obtained by using Ptolemy's inequality:
-
-	\begin{equation*}
-		\begin{aligned}
-			\label{eq4.7}
-			d(a,c') \cdot d(b,c) & \leq d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')                \\
-			d(a,c')              & \leq \frac{d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')}{d(b,c)}
-		\end{aligned}
-	\end{equation*}
-	By replacing some distances with upper boundaries you get the desired upper boundaries.
-\end{proof}
-
-\section{Applying Ptolemy's Inequality to the Elkan's Algorithm}
-
-\subsection{Integration in k-Means}
-Theorem \ref{thm4.2} will now be applied in a similar context to Elkan's algorithm \ref{alg:elkan}, with the goal of integrating these new bounds into steps 5 and 6 of the algorithm.
-
-Consider the data set \(D = \{x_1, \ldots, x_n\}\).
-For each \(x_i\), there is an upper \(u_i\) and lower bounds \(l_{ij}\) as in Elkan's algorithm.
-The new cluster center assigned in a iteration is denoted by \(c_i^{\text{new}}\), and the old cluster center by \(c_i^{\text{old}}\).
-To use Ptolemy's Inequality, one additional point is needed.
-The idea is to use an even older cluster center \(c_i^{\text{old,old}}\) obtained two iterations ago.
-
-Overall, to obtain a new upper bound for the distance between the datapoint \(x_i\) and the new cluster center \(c_j^{\text{new}}\), consider the old cluster center \(c_j^{\text{old}}\) from the last iteration to which \(x_i\) was assigned, and the even older cluster center \(c_j^{\text{old,old}}\) from two iterations ago.
-Utilize the old upper bounds that were valid for the respective iteration.
-
-\begin{equation*}
-	\begin{aligned}
-		d(x_i,c_i^{old})      & \leq u_i^{old}      \\
-		d(x_i,c_i^{old, old}) & \leq u_i^{old, old}
-	\end{aligned}
-\end{equation*}
-
-Similarly, to obtain a new lower bound for the distance between the datapoint \(x_i\) and the new cluster center \(c_j^{\text{new}}\), the old lower bounds from the respective iterations are used:
-
-\begin{equation*}
-	\begin{aligned}
-		l_{i,j}^{\text{old}}     & \leq d(x_i, c_j^{\text{old}})     \\
-		l_{i,j}^{\text{old,old}} & \leq d(x_i, c_j^{\text{old,old}})
-	\end{aligned}
-\end{equation*}
-
-Using Theorem \ref{thm4.2}, the following upper bound for the distance are obtained:
-\begin{equation}
-	\label{eq4.7}
-	d(x_i, c_j^{\text{new}}) \leq u_i^{\text{new}} = \frac{1}{d(c_j^{\text{old}}, c_j^{\text{old,old}})} \cdot \left( u_i^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}) + u_i^{\text{old,old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old}}) \right)
-\end{equation}
-
-and the new lower bound:
-\begin{equation}
-	\label{eq4.8}
-	d(x_i, c_j^{\text{new}}) \geq l_{i,j}^{\text{new}} = \frac{1}{d(c_j^{\text{old}}, c_j^{\text{old,old}})} \cdot \max \left\{
-	\begin{array}{l}
-		l_{i,j}^{\text{old,old}} \cdot d(c_j^{\text{old}}, c_j^{\text{new}}) - u_i^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}), \\
-		l_{i,j}^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}) - u_i^{\text{old,old}} \cdot d(c_j^{\text{old}}, c_j^{\text{new}})
-	\end{array}
-	\right\}
-\end{equation}
-
-You can see that the old upper boundaries are required to calculate the new lower boundaries.
-
-Let us now look at how these bounds can be integrated into Elkan's algorithm.
-It is important to save not only the cluster centres and the bounds of the last iteration, but also those of the penultimate iteration.
-
-At the beginning, all distances to all cluster centres are calculated in the first iteration.
-The upper bound is set by the distance to the nearest cluster centre, and the lower bounds are determined by the direct distances from the point to the respective cluster centres.
-An iteration is then performed as with Elkan, which means that bounds from two iterations are now available.
-This makes it possible to apply the new bounds from equations \ref{eq4.7} and \ref{eq4.8}.
-The algorithm then iterates, as in the original Elkan algorithm using updating with new update conditions, until the cluster centres converge or a specified number of iterations is reached.
-It is always important to check whether $d(c_j^{\text{old}}, c_j^{\text{old,old}}) \neq 0$.
-If this is the case, the update is carried out as for Elkan.
-The algorithm is displayed in Algorithm \ref{alg:ptolemy}.
-
 \begin{algorithm}
-	\caption{Elkan's Algorithm extended with Ptolemy's inequality}
-	\label{alg:ptolemy}
-
+	\caption{Elkan's Algorithm}
+	\label{alg:elkan}
 	\begin{algorithmic}
-
 		\State \textbf{Initialization:} Initialize all cluster centers.
 For each point $x_i$ and each center $c_j$, set the lower bound $l(x_i,c_j)$ and the upper bound $u(x_i)$.
-Assign each $x_i$ to the nearest cluster $C_j$ such that $c(x_i) = \arg \min_j d(x_i, c_j)$, utilizing Lemma 1 to minimize distance calculations.
+Assign each $x_i$ to the nearest cluster $C_j$ such that $c(x_i) = \argmin_j d(x_i, c_j)$, utilizing Lemma 1 to minimize distance calculations.
 Set $r(x_i) = \text{true}$ for all points.
+		\todo{include Lemma 1}
 
-		\State \textbf{Step 0:} Compute one iteration of Elkan's algorithm. Save old upper and lower bounds of the last two iterations.
 		\Repeat
 		\State \textbf{Step 1:} Compute distances $d(c_i, c_j)$ between all centers, and calculate $s(c_i) = \frac{1}{2} \min_{c_j \neq c_i} d(c_i, c_j)$ for each center $c_i$.
-		\Statex
+
 		\State \textbf{Step 2:} Retain points $x_i$ in their current clusters if $u(x_i) \leq s(c(x_i))$.
-		\Statex
+
 		\State \textbf{Step 3:} For remaining points, consider $x_i$ for reassignment if:
-		\begin{itemize}
+		\begin{itemize}[leftmargin=5em,itemindent=\algorithmicindent,itemsep=0pt,parsep=0pt]
 			\item $c_j \neq c(x_i)$,
 			\item $u(x_i) > l(x_i, c_j)$, and
-			\item $u(x_i) > \frac{1}{2}d(c(x_i), c_j)$.
+			\item $u(x_i) > 0.5 \cdot d(c(x_i), c_j)$.
 		\end{itemize}
-		\Statex
-		\State \textbf{Step 3a:}  If $r(x_i)$ is true, compute $d(x_i, c(x_i))$. Set $r(x_i) = \text{false}$. Otherwise, $u(x_i) =d(x_i, c(x_i))$
-		\Statex
+
+		\State \textbf{Step 3a:} If $r(x_i)$ is true, compute $d(x_i, c(x_i))$. Set $r(x_i) = \text{false}$. Otherwise, $u(x_i) = d(x_i, c(x_i))$.
+
 		\State \textbf{Step 3b:} If $d(x_i, c(x_i)) > l(x_i, c_j)$ or $d(x_i, c(x_i)) > \frac{1}{2}d(c(x_i), c_j)$, compute $d(x_i, c_j)$. Reassign $x_i$ to $C_j$ if $d(x_i, c_j) < d(x_i, c(x_i))$.
-		\Statex
+
 		\State \textbf{Step 4:} Compute the cluster centers as the centroids of the corresponding clusters $c'_j$.
-		\Statex
-		\State \textbf{Step 5:} If distance between two older cluster is not 0.
-Update lower bounds by using equation \ref{eq4.8}.
-Otherwise, update like in Elkan.
-Save the old lower bounds of the two last iterations.
-		\Statex
-		\State \textbf{Step 6:} If distance between two older cluster is not 0.
-Update upper bounds by using equation \ref{eq4.7}.
-Otherwise, update like in Elkan.
-Reset $r(x_i) = \text{true}$.
-Save old cluster and upper bounds of the last two iterations.
-		\Statex
+
+		\State \textbf{Step 5:} Update lower bounds $l(x_i, c_j)$ for each $x_i$ and $c_j$ using Eq.~\ref{eq:elkan_lower}.
+		\State \textbf{Step 6:} Update upper bounds $u(x_i)$ for each $x_i$ using Eq.~\ref{eq:elkan_upper}. Reset $r(x_i) = \text{true}$
+
 		\State \textbf{Step 7:} Replace each center $c_j$ with $c'_j$.
-
 		\Until{convergence}
-
 	\end{algorithmic}

### 2024-10-02 content/4_contribution.tex
split methods

--- 2024-10-01 content/3u4_methods.tex
@@
-\section{Foundations of Ptolemaic Indexing}
+\section{Applying Ptolemy's Inequality to the Elkan's Algorithm}
+\label{sec:contrib}
 
@@
 \end{proof}
-
-\section{Applying Ptolemy's Inequality to the Elkan's Algorithm}
 
@@
 The algorithm is displayed in Algorithm \ref{alg:ptolemy}.
-
-\begin{algorithm}
-	\caption{Elkan's Algorithm extended with Ptolemy's inequality}
-	\label{alg:ptolemy}
-
-	\begin{algorithmic}
-
-		\State \textbf{Initialization:} Initialize all cluster centers.
-For each point $x_i$ and each center $c_j$, set the lower bound $l(x_i,c_j)$ and the upper bound $u(x_i)$.
-Assign each $x_i$ to the nearest cluster $C_j$ such that $c(x_i) = \arg \min_j d(x_i, c_j)$, utilizing Lemma 1 to minimize distance calculations.
-Set $r(x_i) = \text{true}$ for all points.
-
-		\State \textbf{Step 0:} Compute one iteration of Elkan's algorithm. Save old upper and lower bounds of the last two iterations.
-		\Repeat
-		\State \textbf{Step 1:}  Compute distances $d(c_i, c_j)$ between all centers, and calculate $s(c_i) = \frac{1}{2} \min_{c_j \neq c_i} d(c_i, c_j)$ for each center $c_i$.
-		\Statex
-		\State \textbf{Step 2:}  Retain points $x_i$ in their current clusters if $u(x_i) \leq s(c(x_i))$.
-		\Statex
-		\State \textbf{Step 3:}  For remaining points, consider $x_i$ for reassignment if:
-		\begin{itemize}
-			\item $c_j \neq c(x_i)$,
-			\item $u(x_i) > l(x_i, c_j)$, and
-			\item $u(x_i) > \frac{1}{2}d(c(x_i), c_j)$.
-		\end{itemize}
-		\Statex
-		\State \textbf{Step 3a:}  If $r(x_i)$ is true, compute $d(x_i, c(x_i))$. Set $r(x_i) = \text{false}$. Otherwise, $u(x_i) =d(x_i, c(x_i))$
-		\Statex
-		\State \textbf{Step 3b:} If $d(x_i, c(x_i)) > l(x_i, c_j)$ or $d(x_i, c(x_i)) > \frac{1}{2}d(c(x_i), c_j)$, compute $d(x_i, c_j)$. Reassign $x_i$ to $C_j$ if $d(x_i, c_j) < d(x_i, c(x_i))$.
-		\Statex
-		\State \textbf{Step 4:} Compute the cluster centers as the centroids of the corresponding clusters $c'_j$.
-		\Statex
-		\State \textbf{Step 5:} If distance between two older cluster is not 0.
-Update lower bounds by using equation \ref{eq4.8}.
-Otherwise, update like in Elkan.
-Save the old lower bounds of the two last iterations.
-		\Statex
-		\State \textbf{Step 6:} If distance between two older cluster is not 0.
-Update upper bounds by using equation \ref{eq4.7}.
-Otherwise, update like in Elkan.
-Reset $r(x_i) = \text{true}$.
-Save old cluster and upper bounds of the last two iterations.
-		\Statex
-		\State \textbf{Step 7:} Replace each center $c_j$ with $c'_j$.
-
-		\Until{convergence}
-
-	\end{algorithmic}
-\end{algorithm}

### 2024-10-02 content/alg_elkan2.tex
fix missing lemma in algorithm

@@
 For each point $x_i$ and each center $c_j$, set the lower bound $l(x_i,c_j)$ and the upper bound $u(x_i)$.
-Assign each $x_i$ to the nearest cluster $C_j$ such that $c(x_i) = \argmin_j d(x_i, c_j)$, utilizing Lemma 1 to minimize distance calculations.
-Set $r(x_i) = \text{true}$ for all points.
-		\todo{include Lemma 1}
+Assign each $x_i$ to the nearest cluster $C_j$ such that $c(x_i) = \argmin_j d(x_i, c_j)$, utilizing that
+		$ d(c_j, c_m) \geq 2d(x_i, c_j) \Rightarrow d(x_i, c_m) \geq d(x_i, c_j) $
+		to minimize distance calculations. Set $r(x_i) = \text{true}$ for all points.
 

### 2024-10-02 content/4_contribution.tex
finally, a full proof

@@
-\section{Applying Ptolemy's Inequality to the Elkan's Algorithm}
+%\newcommand{\coo}{c^{\diamond}}
+\newcommand{\prev}{\diamond}
+
+\section{Contribution: Applying Ptolemy's Inequality to the Elkan's Algorithm}
 \label{sec:contrib}
 
-\todo{explain Hetland's lower (upper) bound notation ($d^-$ and $d^+$)}
-\begin{theorem}[Ptolemaic Bounds]
-	\label{thm4.2}
-	\todo{we might want to explain what the points are supposed to do before showing this theorem.
-		Or, alternatively, we just present this as a purely mathematical statement with no preceding motivation.}
-	Let \(a, b, c, c' \in \mathbb{R}^n\) and let \(d(\cdot, \cdot)\) be the Euclidean distance. The distance \(d(a, c')\) can be bounded as follows:
-	\begin{gather}
-		\frac{1}{d(b, c)} \cdot \max \left\{
+As seen in the previous section,
+Elkan's algorithm can improve the execution speed of $k$-means clustering through the use of the triangle inequality.
+Explicit distance calculations are avoided when tight upper and lower bounds to the distance function are available.
+
+In this section, we use Ptolemy's inequality to calculate an additional set of bounds and modify Elkan's algorithm accordingly.
+These Ptolemaic bounds often improve upon the triangular bounds, leading to increased execution speeds, as will be shown in \autoref{sec:results}.
+
+\subsection{Ptolemaic Bounds}
+Recall that the triangle inequality relates the distances between three points,
+while Ptolemy's inequality does so for four points.
+At the end of an iteration of Elkan's algorithm (Step~5~to~7 in \autoref{alg:elkan}),
+the distance bounds between the data points $x_i$ and the new center positions are calculated $c'_j$.
+Therefore, the three points used in the calculation of the bounds are the aforementioned two points and $c_j$, the cluster center calculated in the previous iteration.
+
+A natural extension to this procedure is then to use an even older cluster center $c^\prev_j$ as a fourth point,
+i.e. the cluster center calculated in the penultimate iteration.
+To formalize, Ptolemy's inequality can be used to calculate the following upper and lower bounds:
+
+\begin{theorem}[Ptolemaic bounds for Elkan's algorithm]
+	During an iteration of Elkan's algorithm,
+	let $x_i \in D \subset \mathbb{R}^n$ be a point from the dataset to be clustered,
+	$d$ the Euclidean distance function,
+	$c'_j$ a cluster center calculated in this iteration,
+	$c_j$ a cluster center calculated in the previous iteration, and
+	$c^\prev_j$ a cluster center calculated in the iteration before the previous iteration.
+	A lower and upper bound on $d(x_i,c'_i)$ is then given by
+	\begin{align}
+		\label{eq:pto_upper}
+		d(x_i, c_j') & \leq u_i' = \frac{1}{d(c_j, c_j^{\prev})} \cdot \left( u_i \cdot d(c_j', c_j^{\prev}) + u_i^{\prev} \cdot d(c_j', c_j) \right) \\
+		\label{eq:pto_lower}
+		d(x_i, c_j') & \geq l_{i,j}' = \frac{1}{d(c_j, c_j^{\prev})} \cdot \max \left\{
 		\begin{array}{l}
-			d(a, b)^- \cdot d(c, c') - d(a, c)^+ \cdot d(b, c'), \\
-			d(a, c)^- \cdot d(b, c') - d(a, b)^+ \cdot d(c, c')
+			l_{i,j}^{\prev} \cdot d(c_j, c_j') - u_i \cdot d(c_j', c_j^{\prev}) \\
+			l_{i,j} \cdot d(c_j', c_j^{\prev}) - u_i^{\prev} \cdot d(c_j, c_j')
 		\end{array}
-		\right\} \leq d(a, c')
-	\end{gather}
-	and
-	\begin{gather}
-		d(a, c') \leq \frac{1}{d(b, c)} \cdot \left( d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c') \right)
-	\end{gather}
+		\right\}\;,
+	\end{align}
+	where $u_i$ and $u^\prev_i$ are the lower bounds calculated in the previous and the penultimate iteration, respectively.
 \end{theorem}
 \begin{proof}
-	\todo{convert to passive voice}
-	\todo{expand context to summarize all steps that lead to this proof}
-	The lower bound corresponds to the derivation in equation \ref{eq4.4} and corresponds to the arrangement of the two pivot objects.
-Replace some distances with the respective lower or upper bound for the distance.
-The lower bounds can be used as they are always smaller than the exact distance.
-The upper bound must be used due to the subtraction.
-This means that the inequality still applies.
-The upper bound is obtained by using Ptolemy's inequality:
+	As all points are in the $\mathrm{R^n}$ and the distances function is also Euclidean, Ptolemy's inequality
+	$d(x, y)\cdot d(v, u) \leq d(x, v) \cdot d(y,v) + d(x, u) \cdot d(y, v)$ holds.
+  Eq.~\ref{eq:pto_uppper} can easily be shown by rearranging Ptolemy's inequality and
+  noting that the upper bounds $u_i, u^\prev_i$ used in place of exact distances can only increase the right-hand side of the inequality. Thus, $d(x_i, c_j') \leq u_i'$ holds.
 
-	\begin{equation*}
-		\begin{aligned}
-			\label{eq4.7}
-			d(a,c') \cdot d(b,c) & \leq d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')                \\
-			d(a,c')              & \leq \frac{d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')}{d(b,c)}
-		\end{aligned}
-	\end{equation*}
-	By replacing some distances with upper boundaries you get the desired upper boundaries.
+  The process is similar for Eq.~\ref{eq:pto_lower}; as the signs on the right hand side are not equal in this case, there are two distinct ways to substitute Ptolemy's inequality.
+  This gives rise to the two inequalities shown inside the maximum function.
+  Analogous to the previous term, the right-hand side of Eq.~\ref{eq:pto_lower} can only be made smaller by inserting lower bounds in the minuend or upper bounds in the subtrahend.
+  Thus, $d(x_i, c_j') & \geq l_{i,j}'$ also holds.
 \end{proof}
 
-\subsection{Integration in k-Means}
-Theorem \ref{thm4.2} will now be applied in a similar context to Elkan's algorithm \ref{alg:elkan}, with the goal of integrating these new bounds into steps 5 and 6 of the algorithm.
+% As Elkan's algorithm tries to avoid distance calculations during the assignment step,
+% the bounds for the clusters–datapoint distances $d(x,c')$ are needed.
+% Thus, two of the point in the triangle inequality are the the data point $x$ and the new cluster position, while the third point 
+% As the majority of potential distance calulations are needed
+% Elkan's algorithm distance computations during The lower and upper bounds 
+% Two of these points have a 
+% In Elkan's original algorithm, the three points
 
-Consider the data set \(D = \{x_1, \ldots, x_n\}\).
-For each \(x_i\), there is an upper \(u_i\) and lower bounds \(l_{ij}\) as in Elkan's algorithm.
-The new cluster center assigned in a iteration is denoted by \(c_i^{\text{new}}\), and the old cluster center by \(c_i^{\text{old}}\).
-To use Ptolemy's Inequality, one additional point is needed.
-The idea is to use an even older cluster center \(c_i^{\text{old,old}}\) obtained two iterations ago.
-
-Overall, to obtain a new upper bound for the distance between the datapoint \(x_i\) and the new cluster center \(c_j^{\text{new}}\), consider the old cluster center \(c_j^{\text{old}}\) from the last iteration to which \(x_i\) was assigned, and the even older cluster center \(c_j^{\text{old,old}}\) from two iterations ago.
-Utilize the old upper bounds that were valid for the respective iteration.
-
-\begin{equation*}
-	\begin{aligned}
-		d(x_i,c_i^{old})      & \leq u_i^{old}      \\
-		d(x_i,c_i^{old, old}) & \leq u_i^{old, old}
-	\end{aligned}
-\end{equation*}
-
-Similarly, to obtain a new lower bound for the distance between the datapoint \(x_i\) and the new cluster center \(c_j^{\text{new}}\), the old lower bounds from the respective iterations are used:
-
-\begin{equation*}
-	\begin{aligned}
-		l_{i,j}^{\text{old}}     & \leq d(x_i, c_j^{\text{old}})     \\
-		l_{i,j}^{\text{old,old}} & \leq d(x_i, c_j^{\text{old,old}})
-	\end{aligned}
-\end{equation*}
-
-Using Theorem \ref{thm4.2}, the following upper bound for the distance are obtained:
-\begin{equation}
-	\label{eq4.7}
-	d(x_i, c_j^{\text{new}}) \leq u_i^{\text{new}} = \frac{1}{d(c_j^{\text{old}}, c_j^{\text{old,old}})} \cdot \left( u_i^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}) + u_i^{\text{old,old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old}}) \right)
-\end{equation}
-
-and the new lower bound:
-\begin{equation}
-	\label{eq4.8}
-	d(x_i, c_j^{\text{new}}) \geq l_{i,j}^{\text{new}} = \frac{1}{d(c_j^{\text{old}}, c_j^{\text{old,old}})} \cdot \max \left\{
-	\begin{array}{l}
-		l_{i,j}^{\text{old,old}} \cdot d(c_j^{\text{old}}, c_j^{\text{new}}) - u_i^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}), \\
-		l_{i,j}^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}) - u_i^{\text{old,old}} \cdot d(c_j^{\text{old}}, c_j^{\text{new}})
-	\end{array}
-	\right\}
-\end{equation}
-
-You can see that the old upper boundaries are required to calculate the new lower boundaries.
-
-Let us now look at how these bounds can be integrated into Elkan's algorithm.
-It is important to save not only the cluster centres and the bounds of the last iteration, but also those of the penultimate iteration.
-
-At the beginning, all distances to all cluster centres are calculated in the first iteration.
-The upper bound is set by the distance to the nearest cluster centre, and the lower bounds are determined by the direct distances from the point to the respective cluster centres.
-An iteration is then performed as with Elkan, which means that bounds from two iterations are now available.
-This makes it possible to apply the new bounds from equations \ref{eq4.7} and \ref{eq4.8}.
-The algorithm then iterates, as in the original Elkan algorithm using updating with new update conditions, until the cluster centres converge or a specified number of iterations is reached.
-It is always important to check whether $d(c_j^{\text{old}}, c_j^{\text{old,old}}) \neq 0$.
-If this is the case, the update is carried out as for Elkan.
-The algorithm is displayed in Algorithm \ref{alg:ptolemy}.
+\subsection{Integration}

### 2024-10-02 content/5_results.tex
finally, a full proof

@@
-\todo{If we claim to have implemented the classic k-means, we also have to show that in our plots.}
-Our variant of the k-Means algorithm is tested by implementing the classical k-Means algorithm (Lloyd's algorithm), Elkan's algorithm, and its novel, Ptolemy-based extension in Python.
-The code is available on GitHub~\cite{Averitchev2024kMeansPtolemy}.
-To provide a fair comparison of the algorithms,
-we measure the number of distance calculations instead of wall-clock time.
-This metric was chosen as distance calculations account for the majority of computing time in k-Means algorithms, offering a clear measure of efficiency regardless of implementational optimizations.
+\section{Preliminary Results}
+\label{sec:results}
 
-Additionally, two intermediate variants were implemented to assess the separate effects of the lower and upper bounds:
-\emph{Ptolemy\_upper} one uses Elkan's lower bound with Ptolemy's upper bound, and \emph{Ptolemy\_lower} uses Ptolemy's lower bound with Elkan's upper bound.
+In order to demonstrate the efficacy of our proposed algorithm, we undertake a comparative analysis with the performance of Elkan's algorithm, which it extends.
+We measure the number of distance evaluations for multiple datasets and for different values of $k$, the number of clusters.
+Our implementation is available through GitHub~\cite{Averitchev2024kMeansPtolemy}.
 
 \subsection{Datasets}
-We employ two groups of datasets to compare the different algorithms, summarized in Table~\ref{tab:datasets}:
-The two real-world datasets include the Iris and Wine datasets \cite{pedregosa2011scikit}, which have fixed dimensions and cardinalities and represent different clustering structures.
-The two synthetic datasets include Gaussian and uniformly randomly generated points, which are used to test for the effects of changing dataset dimensionality.
 
-\begin{table}[bp]
+\begin{table}[b]
 	\centering
-	\caption{Datasets Used in the Experiments}
+	\caption{Datasets used in the experiments}
 	\label{tab:datasets}
@@
 
+We employ two groups of datasets to compare the different algorithms, summarized in Table~\ref{tab:datasets}:
+The two real-world datasets include the Iris and Wine datasets \cite{pedregosa2011scikit}, which have fixed dimensions and cardinalities and represent different clustering structures.
+The two synthetic datasets include Gaussian and uniformly randomly generated points, which are used to test for the effects of changing dataset dimensionality.
+
 \subsection{Experiments}
-The experiments were designed to evaluate the algorithms based on the number of distance calculations, which dominate the computing time in k-Means algorithms.
+\todo{If we claim to have implemented the classic k-means, we also have to show that in our plots.}
+Our variant of the k-Means algorithm is tested by implementing the classical k-Means algorithm (Lloyd's algorithm), Elkan's algorithm, and its novel, Ptolemy-based extension in Python.
+
+To provide a fair comparison of the algorithms,
+we measure the number of distance calculations instead of wall-clock time.
+As distance calculations account for the majority of computing time in k-Means algorithms, they offer a clear measure of efficiency regardless of implementational optimizations.
+
+Additionally, two intermediate variants were implemented to assess the separate effects of the lower and upper bounds:
+\emph{Ptolemy\_upper} one uses Elkan's lower bound with Ptolemy's upper bound, and \emph{Ptolemy\_lower} uses Ptolemy's lower bound with Elkan's upper bound.
+
+Each algorithm is executed three times on each dataset, with the parameters $k = 3$, $k = 20$, and $k = 100$.
+
+In \autoref{fig:combined}, we report our results as distance calculations saved relative to Elkan's algorithm, showcasing any improvements introduced by the proposed extension.
+
+%The experiments evaluate the algorithms based on the number of distance calculations, which dominate the computing time in k-Means algorithms.
 This metric ensures comparability across different implementations and hardware setups.
-Each algorithm was executed three times on each dataset, with the parameters $k = 3$, $k = 20$, and $k = 100$.
-
-The results for the Iris and Wine datasets are visualized in Fig.
-\ref{fig:wine+iris}, while Fig.
-\ref{fig:gauss_univariate} illustrates the performance on the Random and Gaussian dataset variations.
-These figures present the speedup relative to Elkan's algorithm, showcasing any improvements introduced by the proposed extension.
 
@@
 	\includegraphics[width=\textwidth]{fig/combined_plot.pdf}
-	\caption{}
+	\caption{
+		Relative number of saved distance evaluations, compared to Elkan's algorithm.
+		Top and middle row: Performance increases on synthetic data generated from Gaussian (top) and uniformly random (middle) distributions.
+		The dimensionality of the dataset increases left-to-right.\\
+		Bottom row: Performance increases on the real-world datasets Iris (left) and Wine (right).
+	}
 	\label{fig:combined}
@@
 Overall, the full Ptolemaic algorithm proves to be a robust and efficient alternative to traditional k-Means implementations, especially as the complexity of the clustering task increases.
+
+\todo{limitations subsection!}

### 2024-10-02 content/4_contribution.tex
fixing bugs

@@
-%\newcommand{\coo}{c^{\diamond}}
 \newcommand{\prev}{\diamond}
@@
 
-  The process is similar for Eq.~\ref{eq:pto_lower}; as the signs on the right hand side are not equal in this case, there are two distinct ways to substitute Ptolemy's inequality.
-  This gives rise to the two inequalities shown inside the maximum function.
-  Analogous to the previous term, the right-hand side of Eq.~\ref{eq:pto_lower} can only be made smaller by inserting lower bounds in the minuend or upper bounds in the subtrahend.
-  Thus, $d(x_i, c_j') & \geq l_{i,j}'$ also holds.
+	The process is similar for Eq.~\ref{eq:pto_lower}; as the signs on the right-hand side are not equal in this case, there are two distinct ways to substitute into Ptolemy's inequality.
+	This gives rise to the two inequalities, of which the maximum is the stronger bound.
+	Analogous to $u'_i$, the right-hand side of Eq.~\ref{eq:pto_lower} can only be made smaller by inserting lower bounds of the distance $l_{i,j},l_{i,j}^\prev$ in the minuend or upper bounds $u_i, u^\prev_i$ in the subtrahend.
+	Thus, $d(x_i, c_j') \geq l_{i,j}'$ also holds.
 \end{proof}

### 2024-10-02 content/3_preliminaries.tex
merge overlea

\newcommand{\kmeans}{$k$-means problem\xspace}

\section{Preliminaries}\label{Preliminaries}

In this section, we first define the k-means clustering problem and the Lloyd algorithm as a solution to this problem in Section \ref{subsec: kmeans}. We then continue with introducing means of acceleration via lower and upper bounding in Section \ref{sub:acc}.

\subsection{The \kmeans} \label{subsec: kmeans}

To discuss the \kmeans, we first need to formalize the concept of a \emph{clustering algorithm}:
\begin{definition}[clustering algorithm]
	Given a target number of clusters $k$,
	let $F$ be a function that assigns each element from an input data set $D\subset \mathbb{R}^n$ to a
	cluster index $i \in \{1, 2, \ldots, k\}$
	$$ F:D \to \{1, 2, \ldots, k\} \;.$$
\end{definition}
Notably, we restrict our analysis to clustering problems that use the $n$-dimensional Euclidean space as the domain of the input data set.
Likewise, we assume that the similarity of the data points to be clustered is completely defined by the Euclidean distance $d(x,y)= || x-y ||_2 $.
The rationale behind this assumption will be made clear shortly in \autoref{sub:acc}.

The heart of the \kmeans is the assignment of clusters so that the variance in similarity between all points in the same cluster should be reasonably small.
More formally:
\begin{definition}[\kmeans]
	Let $F$ be a clustering algorithm
	%$x \in D$ be an element from the input data set of a clustering algorithm $F$
	and let $C_i$ be the set of all elements assigned to cluster $i$:
	$$ C_i = \{x \in D \mid F(x) = i\} \,.$$
	Furthermore, let $\epsilon$ be the weighted sum of inter-cluster variances
	$$ \epsilon = \sum_{i=1}^k |C_i| \operatorname{Var}[C_i] = \sum_{i=1}^k \sum_{x_j \in C_i}  d^2(x_j, c_i)\,, $$
	where $c_i = \operatorname{E}[C_i]$ is center of cluster $C_i$.

	Then, the map $F$ is called a solution to the \kmeans
	if $\epsilon$ cannot be made smaller by changing the assignment of any one element to a different cluster.
\end{definition}
Note that with this definition, the common $k$-means algorithms provides solution to the \kmeans,
as we do not require that $F$ finds the smallest possible $\epsilon$,
just a local minimum of $\epsilon$.
This is consistent with the colloquial notion that solutions to the \kmeans ``can be found'',
even when an optimal solution (an NP-hard problem) is not practically obtainable \cite{}.

The most prevalent algorithm employed to identify valid solutions to the \kmeans is the Lloyd algorithm.
For convenience, it is reproduced in \autoref{alg:lloyd};
given its widespread use, we will only cover its most significant characteristics as they relate to our analysis.

\begin{algorithm}
	\caption{k-Means Algorithm}
	\label{alg:lloyd}

	\textbf{Input:} \( k \): Number of clusters, \( D \): Dataset containing \( n \) objects

	\textbf{Output:} Assignment of each $x_i \in D$ to a clusters

	\begin{algorithmic}[1]
		\State Initialize \( k \) cluster centers \( \{c_1, c_2, \dots, c_k\} \) arbitrarily from \( D \)
		\Repeat
		\label{algstep:assign}
		\State \textcolor{gray}{// Assign \( x_i \) to the closest cluster}
		\For{each object \( x_i \) in \( D \)}
		\State \( C_j \leftarrow C_j \cup \{x_i\} \) where \( j = \underset{j}{\arg\min} \; d^2(x_i, c_j) \)
		\EndFor
		\State \textcolor{gray}{// Update cluster center \( c_j \)}
		\For{each cluster \( C_j \)}
		\State \( c_j \leftarrow \frac{1}{|C_j|} \sum_{x_i \in C_j} x_i \)
		\EndFor
		\Until{no change in cluster assignments}
	\end{algorithmic}
\end{algorithm}

For each iteration of Lloyds algorithm, the assignment step (\autoref{algstep:assign}) needs to identify the closest cluster center for each element from the input dataset $D$.
To do so, Lloyd's algorithm calculates the distances between each cluster and each element from $D$, requiring $k\cdot|D|$ total distance evaluations per iteration.
These frequent distance evaluations can make up a significant share of the algorithms computational costs, especially for high-dimensional datasets.

\subsection{Accelerating $k$-means}
\label{sub:acc}

As we have seen, Lloyd's algorithm requires numerous evaluations of the distance function.
However, the exact distances do not actually need to be calculated explicitly for every element, as it is sufficient to identify which cluster center is the closest to a given element.
This is often possible from geometric considerations alone, when lower and upper distance bounds are available.
Two properties of the Euclidean space are of particular use here:
The triangle inequality
\begin{align}
	\label{eq:tri}
	d(x,y) \leq d(x,z) + d(z,y)
\end{align}
and Ptolemy's inequality
\begin{align}
	\label{eq:pto}
	d(x, y)\cdot d(v, u) \leq d(x, v) \cdot d(y,v) + d(x, u) \cdot d(y, v)
\end{align}
can provide lower and upper bound on the distance between two points given two (respectively five) other distances.

Numerous algorithms have been developed that exploit the former inequality to avoid explicit distance evaluations. However, to the best of our knowledge, the latter has not been employed for the purposes of accelerating k-means clustering yet.

Therefore, we modify Elkan's algorithm, an existing method that already employs the triangle inequality, to also make use of the Ptolemaic inequality.
We provide a short summary of Elkan's algorithm \cite{DBLP:conf/icml/Elkan03} here and describe our extension in \autoref{sec:contrib}.

Among Elkan's algorithm maintains two kinds of distance bounds each iteration of the algorithm:
\begin{enumerate}[label=\roman*]
	\item $u(x) \geq d(x, c_i) \mid i = F(x)$,
	      upper bounds between data points and their currently assigned centers.
	\item $l(x, c_i) \leq d(x,c_i) \forall i \neq F(x)$,
	      lower bounds between points and all other centers.
\end{enumerate}
Intuitively, when the upper bound is larger than all lower bounds $u(x)\leq l(x,c_i) \forall i\neq F(X)$, the cluster assignment of a data point has not changed.
Explicit distances to other clusters are only required when this inequality does not hold, i.e. when the bounds are not tight enough or when the cluster assignment of the data point has indeed changed.
It is immediately evident that the quality of the bounds has a significant impact on the achievable improvements in execution speed:
The larger (smaller) the lower (upper) bound is, the more likely an explicit distance evaluation can be avoided.

While Elkan's algorithm also employs additional techniques to reduce the number of distance calculations, we want to focus on the calculation of the bounds here;
a complete description of the algorithm is given in \autoref{alg:elkan}, as reproduced from \cite{}.

The lower and upper bounds are calculated at the end of each iteration of Elkan's algorithm and are given by
\begin{align}
	\label{eq:elkan_lower}
	l(x_i, c_j) & = \max \{ l(x_i, c_j) - d(c'_j, c_j), 0 \} \\
	\label{eq:elkan_upper}
	u(x_i)      & = u(x_i) + d(c'(x_i), c(x_i)) \,,
\end{align}
where the prime symbol ($c'_j$) indicates the location of the new cluster centers, while unprimed cluster centers
$c_j$ refer to the centers' positions at the end of the previous iteration.
The given equations for the bounds follow directly from the triangle inequality (Eq.~\ref{eq:tri}).

The calculation of these bounds entails an overhead of $\mathrm{O}(k^2)$ distance calculation per iteration.
However, this additional cost tends to be small compared to the number of distance calculations $\mathrm{O}(k\cdot|D|)$ that can potentially saved,
as  $k^2 \ll |D|$ in most practical applications.

\input{content/alg_elkan2.tex}

### 2024-10-02 content/4_contribution.tex
merge overlea

@@
 	$d(x, y)\cdot d(v, u) \leq d(x, v) \cdot d(y,v) + d(x, u) \cdot d(y, v)$ holds.
-	Eq.~\ref{eq:pto_uppper} can easily be shown by rearranging Ptolemy's inequality and
+	Eq.~\ref{eq:pto_upper} can easily be shown by rearranging Ptolemy's inequality and
 	noting that the upper bounds $u_i, u^\prev_i$ used in place of exact distances can only increase the right-hand side of the inequality. Thus, $d(x_i, c_j') \leq u_i'$ holds.

### 2024-10-02 main.tex
split methods
finally, a full proof
merge overlea

@@
 %%% Das optionale Argument (sofern angegeben) wird für die Kopfzeile verwendet.
-\title[Ein Kurztitel]{Extending K-Means Clustering with Ptolemy’s Inequality}
+\title[Extending K-Means Clustering with Ptolemy’s Inequality]{Extending K-Means Clustering with Ptolemy’s Inequality}
 %% \subtitle{Untertitel / Subtitle} % if needed
-\author[1,2]{Max Pernklau}{firstname1.lastname1@affiliation1.org}{0000-0000-0000-0000}
-\author[2]{Nikita Averitchev}{firstname2.lastname2@affiliation2.org}{0000-0000-0000-0000}
-\author[3]{Christian Beecks}{firstname3.lastname3@affiliation1.org}{0000-0000-0000-0000}
+\author[1]{Max Pernklau}{max.pernklau@fernuni-hagen.de}{0009-0007-5520-4093}
+\author[1]{Nikita Averitchev}{naveritchev@gmail.com}{}
+\author[1]{Christian Beecks}{christian.beecks@fernuni-hagen.de}{0009-0000-9028-629X}
 % \author[1]{Firstname4 Lastname4}{firstname4.lastname4@affiliation1.org}{0000-0000-0000-0000}%
-\affil[1]{Universität 1\\Abteilung\\Straße\\Postleitzahl Ort\\Land}
-\affil[2]{University 2 \\Department\\Address\\Country}
-\affil[3]{University 3\\Department\\Address\\Country}
+\affil[1]{FernUniversität in Hagen\\Chair of Data Science\\ 58084 Hagen\\Germany}
+
 \maketitle
+\begin{abstract}
+	Clustering is a fundamental data analytics operation in the field of unsupervised learning.
+Given a database of unknown structure, clustering aims to discover the inherent structure of the data objects according to similarity, such that similar data objects are grouped together, while dissimilar ones are separated in different groups or clusters.
+	In this study, we focus on partitioning methods, specifically the \emph{k-means clustering} approach, which minimizes the intra-cluster variance.
+	As the standard algorithmic approach to k-means clustering, namely the Lloyd algorithm, is neither efficient nor scalable, various adaptations and modifications have been developed, resulting in the family of fast k-means clustering algorithms.
+In this short paper, we extend the clustering algorithm proposed by Elkan with Ptolemy's inequality to prune superfluous distance calculations.
+It is not our intention with this paper to compete with the state-of-the-art algorithmic solutions for the k-means clustering problem; rather, we seek to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, particularly in terms of reducing computations associated with distance evaluations.
 
-\begin{abstract}
-	Clustering is a fundamental data-analytical operation in the field of unsupervised learning.
-Given a database of unknown structure, clustering aims to discover the inherent structure of the data objects according to similarity, such that similar data objects are grouped together and dissimilar ones are separated in different groups or clusters.
-%The resulting groups, which are denoted as clusters, then represent the clustering structure of the database.
-A
-	Among the various approaches of clustering methods, we focus on partitioning methods, i.e. on the specific problem of \emph{k-means clustering}, which minimizes the intra-cluster variance.
+ % neither tri → pto nor pto → tri is true
+ % which generalizes the triangle inequality,
+
 %where the Lloyd algorithm is a prominent and de facto standard implementation.
-	As the standard algorithmic approach to k-means clustering, namely the Lloyd algorithm, lacks on efficiency and scalability, various adaptations and modifications have been developed yielding to the family of fast k-means clustering algorithms.
-In this short paper, we extend the clustering algorithm proposed by Elkan with Ptolemy's inequality, which generalizes the triangle inequality, to prune unnecessary distance calculations.
-The aim of our short paper is not to compete with the state-of-the-art algorithmic solutions for the k-means clustering problem, but to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations.
+
+%The resulting groups, which are denoted as clusters, then represent the clustering structure of the database. A
 
@@
 
-\todo{Americanize spelling (thesis uses chiefly British terms)}
+%\todo{Americanize spelling (thesis uses chiefly British terms)}
 
@@
 
-With the proliferation of complex data sources and the growing demand for more efficient clustering techniques, numerous \emph{fast k-means algorithms} have been introduced \cite{DBLP:conf/icml/Elkan03,DBLP:conf/sdm/Hamerly10,drake2012accelerated,hamerly2015accelerating,DBLP:conf/icml/NewlingF16,DBLP:conf/icml/DingZSMM15}.
+With the proliferation of complex data sources and the growing demand for more efficient clustering techniques, numerous \emph{fast k-means algorithms} have been introduced \cite{DBLP:conf/icml/Elkan03,DBLP:conf/sdm/Hamerly10,drake2012accelerated,hamerly2015accelerating,DBLP:conf/icml/NewlingF16,DBLP:conf/icml/DingZSMM15,DBLP:conf/sisap/SchubertLF21,DBLP:conf/sisap/LangS23}.
 These methods are designed to achieve exact clustering results with significantly reduced computational cost.
-Moreover, they do not require any form of precomputation, making them directly applicable across a wide range of scenarios.
+Moreover, most of them do not require any form of precomputation, making them directly applicable across a wide range of scenarios.
+The major objective of fast k-means algorithms is to reduce the number of distance evaluations required when assigning data objects to cluster centers and to safely exclude cluster centers from the assignment process.
+For this purpose, distances are approximated via lower and upper bounds.
+
+The Elkan algorithm \cite{DBLP:conf/icml/Elkan03}, as one prominent representative of these algorithms particularly suited for high-dimensional scenarios, maintains one upper bound and multiple lower bounds for each data object in order to apply different pruning criteria for each combination of data object and cluster center.
+While the lower and upper bounds of Elkan's algorithm are directly derived from the triangle inequality, we propose to make use of the Ptolemy inequality instead in order to increase the pruning performance.
+With this extension, we do not aim to ultimately compete with the state of the art in k-means clustering, but to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations.
+We believe that our research findings offer promising directions for future research and new research impulses for data scientists and practitioners alike.
+
+This short paper is structured as follows: Section \ref{Preliminaries} outlines the k-means problem and Elkan's approach of acceleration the algorithmic computation.
+Section \ref{sec:contrib} then proposes how to apply Ptolemy's inequality to the aforementioned algorithm.
+The results of our preliminary performance evaluation are summarized in Section \ref{sec: results} while Section \ref{sec: conclusions} concludes this paper with a short outlook on future work.
 
@@
 
-\newpage
+%\newpage
 
-\section{Related Work}
+%\section{Related Work}
+%\todo{remove this section if we run out of time}
 
-\input{content/3u4_methods}
+\input{content/3_preliminaries}
+\input{content/4_contribution}
 
-\section{Preliminary Results}
+\section{Preliminary Results}\label{sec: results}
 
@@
 
-\section{Conclusions and Future Work}
+\section{Conclusions and Future Work} \label{sec: conclusions}
 
-\clearpage
+This short paper has addressed the problem of efficient data clustering by means of the k-means algorithm.
+More specifically, we have shown how to extend the more efficient clustering algorithm proposed by Elkan with novel distance approximations derived from Ptolemy's inequality.
+The results of our preliminary performance evaluation clearly indicate that our proposal achieves significant improvements in performance compared to Elkan's algorithm.
+%Particularly noteworthy is the increased efficiency with a large number of clusters.
+
+Moreover, the use of the Ptolemy's inequality opens up new avenues for future work.
+It would be valuable to explore adaptations of this technique for other k-means algorithms and optimizations.
+In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality can potentially be leveraged to achieve an enhancement in performance, making these clustering algorithms more practical for broader range of  applications.
+\todo{Moreover, the use of Ptolemy’s inequality opens new avenues for future work.
+Exploring adaptations of this technique for other K-Means algorithms and optimizations could be highly valuable.
+In particular, algorithms that rely on the triangle inequality to prune distance calculations might leverage Ptolemy’s inequality to enhance performance, making these clustering algorithms more practical for a broader range of applications.}
+
+\pagebreak
 %%% Angabe der .bib-Datei (ohne Endung) / State .bib file (im Falle der Nutzung von BibTeX)
 \todo{re-enable citations}
-%\bibliography{references,bibliography_thesis,bibliography_iiwas}
-%% \printbibliography % im Falle der Nutzung von biblatex
+\bibliography{references,bibliography_thesis,bibliography_iiwas}
+%\printbibliography % im Falle der Nutzung von biblatex
 \end{document}

### 2024-10-02 content/4_contribution.tex
proofreading

@@
 
-\section{Contribution: Applying Ptolemy's Inequality to the Elkan's Algorithm}
+\section{Applying Ptolemy's Inequality to the Elkan's Algorithm}
 \label{sec:contrib}
@@
 Recall that the triangle inequality relates the distances between three points,
-while Ptolemy's inequality does so for four points.
-At the end of an iteration of Elkan's algorithm (Step~5~to~7 in \autoref{alg:elkan}),
-the distance bounds between the data points $x_i$ and the new center positions are calculated $c'_j$.
+while Ptolemy's inequality does so for four.
+At the end of an iteration of Elkan's algorithm (Step~5~to~7 in Alg.~\autoref{alg:elkan}),
+the distance bounds between the data points $x_i$ and the new center positions $c'_j$ are calculated.
 Therefore, the three points used in the calculation of the bounds are the aforementioned two points and $c_j$, the cluster center calculated in the previous iteration.
@@
 
+\subsection{Integration}
+To integrate the novel bounds into the existing framework of Elkan's algorithm,
+only Steps~6 and~7 in Alg.~\autoref{alg:elkan} need to be updated to use the new bounds.
+It is possible to only replace either the upper or lower bound, which leads to a hybrid solution that is further explored in \autoref{sec: results}.
+
+As the Ptolemaic bounds require bounds from previous iterations, the first iteration of the algorithm is always conducted with Elkan's original bounds.
+
 % As Elkan's algorithm tries to avoid distance calculations during the assignment step,
@@
 % In Elkan's original algorithm, the three points
-
-\subsection{Integration}

### 2024-10-02 main.tex
proofreading

@@
 %intro
-Clustering is a fundamental data-analytical operation in the domain of data science, where the basic task is to divide a set of data objects into different groups or clusters, such that objects within the same cluster have sufficient similarities, while objects in different clusters have significant differences.
+Clustering is a fundamental data-analytical operation in the domain of data science.
+The basic task is to divide a set of data objects into different groups or clusters, such that objects within the same cluster have sufficient similarities;
+meanwhile objects in different clusters should have significant differences.
 The collection of clusters reflects the inherent structure of the database and is denoted as clustering.
@@
 
-The Elkan algorithm \cite{DBLP:conf/icml/Elkan03}, as one prominent representative of these algorithms particularly suited for high-dimensional scenarios, maintains one upper bound and multiple lower bounds for each data object in order to apply different pruning criteria for each combination of data object and cluster center.
-While the lower and upper bounds of Elkan's algorithm are directly derived from the triangle inequality, we propose to make use of the Ptolemy inequality instead in order to increase the pruning performance.
-With this extension, we do not aim to ultimately compete with the state of the art in k-means clustering, but to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations.
+The Elkan algorithm \cite{DBLP:conf/icml/Elkan03} is a prominent representative of these algorithms and particularly suited for high-dimensional scenarios.
+It maintains one upper bound and multiple lower bounds for each data object in order to apply different pruning criteria for each combination of data object and cluster center.
+While the lower and upper bounds of Elkan's algorithm are directly derived from the triangle inequality, we propose to utilitze Ptolemy's inequality instead in order to increase the pruning performance.
+With this extension, we do not aim to compete with the state of the art in k-means clustering;
+rather we intend to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations.
 We believe that our research findings offer promising directions for future research and new research impulses for data scientists and practitioners alike.
 
-This short paper is structured as follows: Section \ref{Preliminaries} outlines the k-means problem and Elkan's approach of acceleration the algorithmic computation.
-Section \ref{sec:contrib} then proposes how to apply Ptolemy's inequality to the aforementioned algorithm.
-The results of our preliminary performance evaluation are summarized in Section \ref{sec: results} while Section \ref{sec: conclusions} concludes this paper with a short outlook on future work.
+This paper is structured as follows:
+Section~\ref{Preliminaries} outlines the k-means problem and Elkan's approach of acceleration the algorithmic computation.
+In Section~\ref{sec:contrib}, we proposes how to apply Ptolemy's inequality to the aforementioned algorithm as our main contribution.
+The results of our preliminary performance evaluation are detailed in Section~\ref{sec: results}, while Section~\ref{sec: conclusions} concludes this paper with a short outlook on future work.
 
@@
 	Experiments to show:
-	- iris, wine, gaussian, uniform
+	- iris, wine, Gaussian, uniform
 	(moons, circles aren't a good fit for k-means -- is that reason enough to exclude them?)
@@
 
-\pagebreak
-%%% Angabe der .bib-Datei (ohne Endung) / State .bib file (im Falle der Nutzung von BibTeX)
-\todo{re-enable citations}
-\bibliography{references,bibliography_thesis,bibliography_iiwas}
-%\printbibliography % im Falle der Nutzung von biblatex
+%\bibliography{references,bibliography_thesis,bibliography_iiwas}
 \end{document}

### 2024-10-02 content/5_results.tex
merge overlea
finally doctored limitations and ending into something readbale

@@
-\section{Preliminary Results}
-\label{sec:results}
-
 In order to demonstrate the efficacy of our proposed algorithm, we undertake a comparative analysis with the performance of Elkan's algorithm, which it extends.
@@
 
-\begin{table}[b]
+\begin{table}[t]
 	\centering
@@
 
-\begin{figure}
+\subsection{Discussion}
+
+\begin{figure}[h]
 	\centering
@@
 
-\subsection{Analysis of the Results}
+The primary focus of this analysis is the full Ptolemaic algorithm, as it is the most robust when comparison to Elkan's algorithm.
+In general, the full Ptolemaic extension shows substantial improvements in reducing the number of distance evaluations across most scenarios.
 
-The experimental results, presented in the figures, provide critical insights into the performance of the k-Means algorithm variants.
+\textbf{Number of Clusters.}
+The full Ptolemaic algorithm consistently demonstrates significant improvements over Elkan’s algorithm as the number of clusters $k$ increases.
+For instance, in the Gaussian low-dimensional dataset, the Ptolemaic algorithm achieves larger than 2-fold speedups for $k = 20$ and $k = 100$, illustrating its ability to significantly reduce distance calculations in complex clustering tasks.
+% surpasses 3-fold improvements for
+% Similar trends are observed in the Random low-dimensional dataset.
+For smaller cluster counts ($k = 3$), the advantage is less pronounced, with speedups around 1.1 in the Iris and Wine datasets.
+This suggests that Ptolemy's inequality offers significant improvements for larger values of $k$.
+However, lower-complexity clustering tasks may not benefit to the same extent,
+given the additional overhead imposed by the Ptolemaic bounds, which are computationally more complex to evaluate compared to the triangle bounds.
+The Ptolemaic Upper Bound variant performs comparably to the full Ptolemaic algorithm, especially for larger $k$ values.
+However, the Lower Bound variant underperforms in some cases, particularly with smaller cluster counts or in higher-dimensional datasets.
+
+\textbf{Dimensionality.} The full Ptolemaic algorithm performs well across varying dimensionality.
+In low-dimensional datasets (10 dimensions), it achieves impressive speedups, particularly at higher cluster counts.
+As dimensionality increases, performance gains persist but become slightly less pronounced.
+Nonetheless, even in higher-dimensional datasets, the full Ptolemaic algorithm remains effective, consistently reducing distance evaluations.
+However, the Ptolemaic lower bound variant often struggles in these higher-dimensional cases, further emphasizing the robustness of the full Ptolemaic approach.
+
+% \textbf{Summary.}
+% The full Ptolemaic algorithm demonstrates consistent and significant improvements over Elkan’s algorithm across a variety of datasets.
+% In particularly at a high number of clusters and low to medium dimensionality.
+% The Ptolemaic upper bound variant performs similarly in many cases, offering comparable efficiency gains, whereas the Ptolemaic lower bound variant, particularly in datasets with higher dimensionality or smaller cluster counts.
+Overall, the full Ptolemaic algorithm proves to be a robust and efficient alternative to traditional k-Means implementations, especially as the complexity of the clustering task increases.
+
+\textbf{Limitations.}
+The presented experiments are subject to three primary constraints:
+(i) While synthetic random datasets provide valuable best-case scenarios, more extensive real-world data with appropriate grouping structures is necessary for comprehensive evaluation.
+(ii) The significant reduction in distance calculations may not directly translate to improved execution times, as Ptolemy's bounds are computationally more expensive than the triangle inequality.
+(iii) Most critically, the limited sample size and fixed initial conditions, particularly the lack of variation in cluster initializations, severely restrict the generalizability of individual measurements.\\
+Despite these limitations, we contend that the observed overall trend remains compelling and warrants further investigation into the application of Ptolemy's inequality for clustering algorithms.
+
+% The experiments presented suffer from two main factors:
+% - minor caveat: synthetic random datasets are good worst-case scenarios, but larger amounts of real-world data is needed.
+% - major caveat: sample size and initial conditions.
+experiments don't vary cluster initializations.
+while the results presented aren't cherry-picked, single measurements contain only little information.
+% - despite these shortcommings, we believe that the overall trend is still convincing.
+
+% The experimental results, presented in the figures, provide critical insights into the performance of the k-Means algorithm variants.
 The primary focus of this analysis is on the full Ptolemaic algorithm and its comparison with Elkan’s algorithm.
 The results highlight several scenarios in which the full Ptolemaic algorithm achieves substantial improvements over Elkan’s approach, particularly in reducing the number of distance evaluations.
-
-\textbf{Influence of the Number of Clusters}
-
-The full Ptolemaic algorithm consistently demonstrates notable improvements over Elkan’s algorithm as the number of clusters ($k$) increases.
+%
+% \textbf{Number of Clusters.} The full Ptolemaic algorithm consistently demonstrates notable improvements over Elkan’s algorithm as the number of clusters ($k$) increases.
 For instance, in the Gaussian low-dimensional dataset, the Ptolemaic algorithm achieves speedups of over 2-fold for $k = 20$ and surpasses 3-fold for $k = 100$, illustrating its ability to significantly reduce distance calculations in more complex clustering tasks.
@@
 These results underscore the increasing efficiency of the Ptolemaic algorithm as the clustering complexity grows.
-
-In contrast, for smaller cluster counts, such as $k = 3$, the performance advantage of the full Ptolemaic algorithm is less pronounced but still evident, with speedups of around 1.1 in both the Iris and Wine datasets.
+%
+% In contrast, for smaller cluster counts, such as $k = 3$, the performance advantage of the full Ptolemaic algorithm is less pronounced but still evident, with speedups of around 1.1 in both the Iris and Wine datasets.
 This suggests that while Ptolemy’s inequality offers significant improvements for larger $k$ values, it still provides moderate gains for smaller cluster sizes.
-
-The Ptolemaic Upper Bound variant performs comparably to the full Ptolemaic algorithm, especially for larger $k$ values, delivering similar efficiency improvements.
+%
+% The Ptolemaic Upper Bound variant performs comparably to the full Ptolemaic algorithm, especially for larger $k$ values, delivering similar efficiency improvements.
 However, the Ptolemaic Lower Bound variant tends to underperform in some cases, particularly with smaller cluster counts or in higher-dimensional datasets, where its speedup is sometimes close to or even below that of Elkan’s algorithm.
-
-\textbf{Influence of Dimensionality}
-
-The full Ptolemaic algorithm also performs well across datasets with varying dimensionality.
+%
+% \textbf{Dimensionality.} The full Ptolemaic algorithm also performs well across datasets with varying dimensionality.
 In the Gaussian low-dimensional dataset (10 dimensions), the full Ptolemaic algorithm achieves impressive speedups, particularly at higher cluster counts, with speedups over 3-fold at $k = 100$.
@@
 For example, in the Gaussian medium (100 dimensions) and Gaussian high (300 dimensions) datasets, the Ptolemaic algorithm continues to offer significant speedups, but the performance advantage over Elkan’s algorithm decreases slightly.
-
-In higher-dimensional datasets, such as the Random high-dimensional dataset, the full Ptolemaic algorithm remains effective, consistently reducing the number of distance evaluations, particularly for larger $k$ values.
+%
+% In higher-dimensional datasets, such as the Random high-dimensional dataset, the full Ptolemaic algorithm remains effective, consistently reducing the number of distance evaluations, particularly for larger $k$ values.
 However, the Ptolemaic Lower Bound variant often struggles in these higher-dimensional cases, further emphasizing the robustness of the full Ptolemaic approach.
-
-\textbf{Summary of Results}
-
-The full Ptolemaic algorithm demonstrates consistent and significant improvements over Elkan’s algorithm across a variety of datasets, particularly in scenarios with a high number of clusters and low to medium dimensionality.
+%
+% \textbf{Summary.} The full Ptolemaic algorithm demonstrates consistent and significant improvements over Elkan’s algorithm across a variety of datasets, particularly in scenarios with a high number of clusters and low to medium dimensionality.
 The Ptolemaic Upper Bound variant performs similarly in many cases, offering comparable efficiency gains, whereas the Ptolemaic Lower Bound variant sometimes struggles, particularly in datasets with higher dimensionality or smaller cluster counts.
 Overall, the full Ptolemaic algorithm proves to be a robust and efficient alternative to traditional k-Means implementations, especially as the complexity of the clustering task increases.
-
-\todo{limitations subsection!}

### 2024-10-02 main.tex
finally doctored limitations and ending into something readbale

@@
 %%% Das optionale Argument (sofern angegeben) wird für die Kopfzeile verwendet.
-\title[Extending K-Means Clustering with Ptolemy’s Inequality]{Extending K-Means Clustering with Ptolemy’s Inequality}
+\title[Extending K-Means Clustering with Ptolemy’s Inequality]{Extending K-Means Clustering with Ptolemy’s Inequality
+\thanks{This work is partly funded by DFG grant No. 454630593: EPIX – Efficient Ptolemaic Indexing.} 
+}
 %% \subtitle{Untertitel / Subtitle} % if needed
@@
 
-\todo{
-	Experiments to show:
-	- iris, wine, Gaussian, uniform
-	(moons, circles aren't a good fit for k-means -- is that reason enough to exclude them?)
-}
-
-\todo{we should mention that the Pto\_low version performs worse in every test (but one), but we do not need to include it in the comparison between Elkan and Our Work}
-
-\todo{compile a graph that summarizes the four pages of tables}
-
-\todo{How much do we "trust" these results, provided that each test configuration ( [k choice]x[dataset]x[algorithm]) was only run once?
-We might need to repeat and average over these experiments.
-TBD
-}
-
 \section{Conclusions and Future Work} \label{sec: conclusions}
 
+% This paper addresses efficient data clustering using the $k$-means algorithm.
+% We extend Elkan's algorithm with novel distance approximations derived from Ptolemy's inequality.
+% Preliminary evaluations show significant performance improvements over Elkan's method, particularly with large numbers of clusters at low-to-medium dimensionalities.
+% The application of Ptolemy's inequality opens new avenues for future work, potentially enhancing other k-means algorithms that use triangle inequality for distance pruning.
+This approach could make clustering algorithms more practical across a broader range of applications.
+
 This short paper has addressed the problem of efficient data clustering by means of the k-means algorithm.
-More specifically, we have shown how to extend the more efficient clustering algorithm proposed by Elkan with novel distance approximations derived from Ptolemy's inequality.
-The results of our preliminary performance evaluation clearly indicate that our proposal achieves significant improvements in performance compared to Elkan's algorithm.
+We have shown how to extend the more efficient clustering algorithm proposed by Elkan with novel distance approximations derived from Ptolemy's inequality.
+The results of our preliminary performance evaluation indicate that our proposal achieves significant improvements in performance compared to Elkan's algorithm.
 %Particularly noteworthy is the increased efficiency with a large number of clusters.
@@
 Moreover, the use of the Ptolemy's inequality opens up new avenues for future work.
-It would be valuable to explore adaptations of this technique for other k-means algorithms and optimizations.
+Besides conducting more extensive experiments to verify our findings,
+it would be valuable to explore adaptations of this technique for other k-means algorithms and optimizations.
 In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality can potentially be leveraged to achieve an enhancement in performance, making these clustering algorithms more practical for broader range of applications.
-\todo{Moreover, the use of Ptolemy’s inequality opens new avenues for future work.
+
+% \todo{Moreover, the use of Ptolemy’s inequality opens new avenues for future work.
 Exploring adaptations of this technique for other K-Means algorithms and optimizations could be highly valuable.
@@
 
-%\bibliography{references,bibliography_thesis,bibliography_iiwas}
+\bibliography{references,bibliography_thesis,bibliography_iiwas}
 \end{document}

### 2024-11-14 content/3_preliminaries.tex
proofreading
unclear changes

@@
 
-To discuss the \kmeans, we first need to formalize the concept of a \emph{clustering algorithm}:
+In order to discuss the \kmeans, it is sensible to define the concept of a \emph{clustering algorithm} formally.
 \begin{definition}[clustering algorithm]
@@
 This is consistent with the colloquial notion that solutions to the \kmeans ``can be found'',
-even when an optimal solution (an NP-hard problem) is not practically obtainable \cite{}.
+even when an optimal solution (an NP-hard problem) is not practically obtainable \cite{han2012data}.
 
 The most prevalent algorithm employed to identify valid solutions to the \kmeans is the Lloyd algorithm.
-For convenience, it is reproduced in \autoref{alg:lloyd};
+For convenience, it is reproduced in Algorithm~\ref{alg:lloyd};
 given its widespread use, we will only cover its most significant characteristics as they relate to our analysis.
 
-\begin{algorithm}
+\begin{algorithm}[t]
 	\caption{k-Means Algorithm}
@@
 		\Repeat
-		\label{algstep:assign}
 		\State \textcolor{gray}{// Assign \( x_i \) to the closest cluster}
 		\For{each object \( x_i \) in \( D \)} 
+		\label{algstep:assign}
 		\State \( C_j \leftarrow C_j \cup \{x_i\} \) where \( j = \underset{j}{\arg\min} \; d^2(x_i, c_j) \)
@@
 \label{sub:acc}
+
+\input{content/alg_elkan2.tex}
 
@@
 While Elkan's algorithm also employs additional techniques to reduce the number of distance calculations, we want to focus on the calculation of the bounds here;
-a complete description of the algorithm is given in \autoref{alg:elkan}, as reproduced from \cite{}.
+a complete description of the algorithm is given in Algorithm~\ref{alg:elkan}, as reproduced from \cite{elkan}.
 
@@
 as  $k^2 \ll |D|$ in most practical applications.
-
-\input{content/alg_elkan2.tex}

### 2024-11-14 content/4_contribution.tex
unclear changes

@@
 while Ptolemy's inequality does so for four.
-At the end of an iteration of Elkan's algorithm (Step~5~to~7 in Alg.~\autoref{alg:elkan}),
+At the end of an iteration of Elkan's algorithm (Step~5~to~7 in Alg.~\ref{alg:elkan}),
 the distance bounds between the data points $x_i$ and the new center positions $c'_j$ are calculated.

### 2024-11-14 content/3_preliminaries.tex
overleaf version as submitted

@@
 
-In this section, we first define the k-means clustering problem and the Lloyd algorithm as a solution to this problem in Section \ref{subsec: kmeans}.
-We then continue with introducing means of acceleration via lower and upper bounding in Section \ref{sub:acc}.
+In this section, we first define the $k$-means clustering problem and the Lloyd algorithm as a straightforward, but optimizable solution to this problem in Section \ref{subsec: kmeans}.
+We then continue with the introduction of means of acceleration trough lower and upper bounding in Section \ref{sub:acc}.
 
@@
 
-In order to discuss the \kmeans, it is sensible to define the concept of a \emph{clustering algorithm} formally.
+In order to discuss the \kmeans, we first define the concept of a \emph{clustering algorithm} in a formal and implementation-agnostic manner:
 \begin{definition}[clustering algorithm]
-	Given a target number of clusters $k$,
-	let $F$ be a function that assigns each element from an input data set $D\subset \mathbb{R}^n$ to a
+	Given a target number of clusters $k \in \mathbb{N}$, a clustering algorithm is a function $F$ that assigns each element from an input data set $D\subset \mathbb{R}^n$ a
 	cluster index $i \in \{1, 2, \ldots, k\}$
@@
 \end{definition}
+The intention is to assign similar objects the same index, thus clustering them together.
 Notably, we restrict our analysis to clustering problems that use the $n$-dimensional Euclidean space as the domain of the input data set.
-Likewise, we assume that the similarity of the data points to be clustered is completely defined by the Euclidean distance $d(x,y)= || x-y ||_2 $.
-The rationale behind this assumption will be made clear shortly in \autoref{sub:acc}.
+Likewise, we assume that the dissimilarity of the data points to be clustered is completely described by the Euclidean distance $d(x,y)= || x-y ||_2 $.
+The rationale behind this assumption will be made clear shortly in Section \ref{sub:acc}.
 
-The heart of the \kmeans is the assignment of clusters so that the variance in similarity between all points in the same cluster should be reasonably small.
+The core of the \kmeans, then, is the assignment of clusters so that the variance in dissimilarity between all points in the same cluster should be reasonably small.
 More formally:
@@
 	%$x \in D$ be an element from the input data set of a clustering algorithm $F$
-	and let $C_i$ be the set of all elements assigned to cluster $i$:
+	and $C_i$ be the set of all elements assigned to cluster $i$:
 	$$ C_i = \{x \in D \mid F(x) = i\} \,.$$
@@
 	$$ \epsilon = \sum_{i=1}^k |C_i| \operatorname{Var}[C_i] = \sum_{i=1}^k \sum_{x_j \in C_i}  d^2(x_j, c_i)\,, $$
-	where $c_i = \operatorname{E}[C_i]$ is center of cluster $C_i$.
+	where the cluster's average element $c_i = \operatorname{E}[C_i]$ is called the center of the cluster $C_i$.
 
 	Then, the map $F$ is called a solution to the \kmeans
-	if $\epsilon$ cannot be made smaller by changing the assignment of any one element to a different cluster.
+	iff $\epsilon$ cannot be made smaller by changing the assignment of any one element to a different cluster.
 \end{definition}
-Note that with this definition, the common $k$-means algorithms provides solution to the \kmeans,
-as we do not require that $F$ finds the smallest possible $\epsilon$,
+Note that with this definition, common $k$-means algorithms provide solutions to this \kmeans,
+as we do not require that $F$ achieves the smallest possible $\epsilon$,
 just a local minimum of $\epsilon$.
@@
 
-The most prevalent algorithm employed to identify valid solutions to the \kmeans is the Lloyd algorithm.
-For convenience, it is reproduced in Algorithm~\ref{alg:lloyd};
+The most prevalent algorithm employed to identify valid solutions to the \kmeans is the Lloyd algorithm \cite{DBLP:journals/tit/Lloyd82}, which is reproduced in Algorithm~\ref{alg:lloyd};
 given its widespread use, we will only cover its most significant characteristics as they relate to our analysis.
@@
 
-For each iteration of Lloyds algorithm, the assignment step (\autoref{algstep:assign}) needs to identify the closest cluster center for each element from the input dataset $D$.
+For each iteration of Lloyd's algorithm, the assignment step (line~\autoref{algstep:assign}) needs to identify the closest cluster center $c_j$ for each element $x_i$ from the input dataset $D$.
 To do so, Lloyd's algorithm calculates the distances between each cluster and each element from $D$, requiring $k\cdot|D|$ total distance evaluations per iteration.
@@
 	\label{eq:pto}
-	d(x, y)\cdot d(v, u) \leq d(x, v) \cdot d(y,v) + d(x, u) \cdot d(y, v)
+	d(x, y)\cdot d(v, u) \leq d(x, v) \cdot d(y,v) + d(x, u) \cdot d(y, v)\;.
 \end{align}
-can provide lower and upper bound on the distance between two points given two (respectively five) other distances.
+Both inequalities, known since ancient times, provide lower and upper bounds on the distances between two points given two (respectively five) other distances.
 
 Numerous algorithms have been developed that exploit the former inequality to avoid explicit distance evaluations.
-However, to the best of our knowledge, the latter has not been employed for the purposes of accelerating k-means clustering yet.
+However, to the best of our knowledge, the latter has not been employed for the purposes of accelerating $k$-means clustering yet.
 
-Therefore, we modify Elkan's algorithm, an existing method that already employs the triangle inequality, to also make use of the Ptolemaic inequality.
-We provide a short summary of Elkan's algorithm \cite{DBLP:conf/icml/Elkan03} here and describe our extension in \autoref{sec:contrib}.
+To demonstrate the potential of the latter, we modify Elkan's algorithm \cite{DBLP:conf/icml/Elkan03} that already employs the triangle inequality, to also make use of Ptolemy's inequality.
+We provide a short summary of Elkan's algorithm here and describe our extension in Section \ref{sec:contrib}.
 
-Among Elkan's algorithm maintains two kinds of distance bounds each iteration of the algorithm:
+Elkan's algorithm maintains two kinds of distance bounds in each iteration of the algorithm:
 \begin{enumerate}[label=\roman*]
@@
 \end{enumerate}
-Intuitively, when the upper bound is larger than all lower bounds $u(x)\leq l(x,c_i) \forall i\neq F(X)$, the cluster assignment of a data point has not changed.
+Intuitively, when the upper bound is larger than all lower bounds $u(x)\geq l(x,c_i) \forall i\neq F(x)$, the cluster assignment of a data point has not changed.
 Explicit distances to other clusters are only required when this inequality does not hold, i.e.
@@
 While Elkan's algorithm also employs additional techniques to reduce the number of distance calculations, we want to focus on the calculation of the bounds here;
-a complete description of the algorithm is given in Algorithm~\ref{alg:elkan}, as reproduced from \cite{elkan}.
+a complete description of the algorithm is given in Algorithm~\ref{alg:elkan}, as reproduced from \cite{DBLP:conf/icml/Elkan03}.
 
@@
 	\label{eq:elkan_lower}
-	l(x_i, c_j) & = \max \{ l(x_i, c_j) - d(c'_j, c_j), 0 \} \\
+	l'(x_i, c_j) & = \max \{ l(x_i, c_j) - d(c'_j, c_j), 0 \} \\
 	\label{eq:elkan_upper}
-	u(x_i)      & = u(x_i) + d(c'(x_i), c(x_i)) \,,
+	u'(x_i)      & = u(x_i) + d(c'(x_i), c(x_i)) \,,
 \end{align}
-where the prime symbol ($c'_j$) indicates the location of the new cluster centers, while unprimed cluster centers
-$c_j$ refer to the centers' positions at the end of the previous iteration.
-The given equations for the bounds follow directly from the triangle inequality (Eq.~\ref{eq:tri}).
+where the prime symbol ($c'_j, u', l'$) indicates quantities calculated in this iteration, while unprimed variables refer to quantities calculated in the preceding iteration.
+The given equations for the bounds follow directly from the triangle inequality~\ref{eq:tri}.
 
-The calculation of these bounds entails an overhead of $\mathrm{O}(k^2)$ distance calculation per iteration.
+The calculation of these bounds entails an overhead of $\mathrm{O}(k^2)$ distance calculations per iteration.
 However, this additional cost tends to be small compared to the number of distance calculations $\mathrm{O}(k\cdot|D|)$ that can potentially saved,
-as  $k^2 \ll |D|$ in most practical applications.
+as  $k \ll |D|$ in most practical applications.

### 2024-11-14 content/5_results.tex
overleaf version as submitted

@@
 In order to demonstrate the efficacy of our proposed algorithm, we undertake a comparative analysis with the performance of Elkan's algorithm, which it extends.
-We measure the number of distance evaluations for multiple datasets and for different values of $k$, the number of clusters.
+We measure the number of distance evaluations for different benchmark datasets and for different values of $k$, the number of clusters.
 Our implementation is available through GitHub~\cite{Averitchev2024kMeansPtolemy}.
@@
 	\centering
-	\caption{Datasets used in the experiments}
+	\caption{Datasets used in the experiments.}
 	\label{tab:datasets}
@@
 Additionally, two intermediate variants were implemented to assess the separate effects of the lower and upper bounds:
-\emph{Ptolemy\_upper} one uses Elkan's lower bound with Ptolemy's upper bound, and \emph{Ptolemy\_lower} uses Ptolemy's lower bound with Elkan's upper bound.
+The \emph{Ptolemy Upper Bound} method uses Elkan's lower bound and Ptolemy's upper bound, while \emph{Ptolemy Lower Bound} uses Ptolemy's lower bound and Elkan's upper bound.
 
@@
 
-In \autoref{fig:combined}, we report our results as distance calculations saved relative to Elkan's algorithm, showcasing any improvements introduced by the proposed extension.
+In \autoref{fig:combined},
+we report our results in the form of speedups,
+i.e. as the number of distances calculated by our extension,
+divided by the number of distances calculated by Elkan's algorithm.
 
@@
 	\caption{
-		Relative number of saved distance evaluations, compared to Elkan's algorithm.
-		Top and middle row: Performance increases on synthetic data generated from Gaussian (top) and uniformly random (middle) distributions.
-		The dimensionality of the dataset increases left-to-right.\\
-		Bottom row: Performance increases on the real-world datasets Iris (left) and Wine (right).
+		Relative improvement in the number distance evaluations, compared to Elkan's algorithm.
+        A speedup of $>1$ indicates an improvement over Elkan's original algorithm.
+		\textbf{Top and middle row:} Performance increases on synthetic data generated from Gaussian (top) and uniformly random (middle) distributions.
+		The dimensionality of the dataset increases left-to-right.
+		\textbf{Bottom row:} Performance increases on the real-world datasets Iris (left) and Wine (right).
 	}
@@
 
-The primary focus of this analysis is the full Ptolemaic algorithm, as it is the most robust when comparison to Elkan's algorithm.
+The primary focus of this analysis is the full Ptolemaic algorithm as compared to Elkan's algorithm, since it is the most robust one of the three variants. 
 In general, the full Ptolemaic extension shows substantial improvements in reducing the number of distance evaluations across most scenarios.
@@
 
-\textbf{Dimensionality.} The full Ptolemaic algorithm performs well across varying dimensionality.
+\textbf{Dimensionality.} The full Ptolemaic algorithm performs well across varying dataset dimensionality.
 In low-dimensional datasets (10 dimensions), it achieves impressive speedups, particularly at higher cluster counts.
@@
 (ii) The significant reduction in distance calculations may not directly translate to improved execution times, as Ptolemy's bounds are computationally more expensive than the triangle inequality.
-(iii) Most critically, the limited sample size and fixed initial conditions, particularly the lack of variation in cluster initializations, severely restrict the generalizability of individual measurements.\\
-Despite these limitations, we contend that the observed overall trend remains compelling and warrants further investigation into the application of Ptolemy's inequality for clustering algorithms.
+(iii) Most critically, the limited sample size and fixed initial conditions, particularly the lack of variation in cluster initializations, severely restrict the generalizability of individual measurements.
+
+Despite these limitations, we conclude that the observed overall trend remains compelling and warrants further investigation into the application of Ptolemy's inequality for clustering algorithms.
 

### 2024-11-14 main.tex
overleaf version as submitted

@@
 	In this study, we focus on partitioning methods, specifically the \emph{k-means clustering} approach, which minimizes the intra-cluster variance.
-	As the standard algorithmic approach to k-means clustering, namely the Lloyd algorithm, is neither efficient nor scalable, various adaptations and modifications have been developed, resulting in the family of fast k-means clustering algorithms.
+	As the standard algorithmic approach to k-means clustering, the Lloyd algorithm, is neither efficient nor scalable, various adaptations and modifications have been developed, resulting in the family of fast k-means clustering algorithms.\\
     In this short paper, we extend the clustering algorithm proposed by Elkan with Ptolemy's inequality to prune superfluous distance calculations.
-It is not our intention with this paper to compete with the state-of-the-art algorithmic solutions for the k-means clustering problem; rather, we seek to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, particularly in terms of reducing computations associated with distance evaluations.
+It is not our intention to compete with the state-of-the-art algorithmic solutions for the k-means clustering problem; rather, we seek to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, particularly in terms of reducing computations associated with distance evaluations.
 
@@
 %intro
-Clustering is a fundamental data-analytical operation in the domain of data science.
+Clustering is a fundamental data-analytics operation in the domain of data science.
 The basic task is to divide a set of data objects into different groups or clusters, such that objects within the same cluster have sufficient similarities;
 meanwhile objects in different clusters should have significant differences.
-The collection of clusters reflects the inherent structure of the database and is denoted as clustering.
+The collection of clusters reflects the inherent structure of the underlying data-generating process and is denoted as a clustering.
 
-Clustering can be found in various scientific or economic application domains \cite{ezugwu2022comprehensive,oyewole2023data, gan2020data}, ranging from archaeology \cite{troiano2024comparative} and finance \cite{cai2016clustering} to industry \cite{lee2021technological} and zoology \cite{shen2021multivariate}, to name just a few.
+The need for clustering arises in various scientific or economic application domains \cite{ezugwu2022comprehensive,oyewole2023data, gan2020data}, ranging from archaeology \cite{troiano2024comparative} and finance \cite{cai2016clustering} to industry \cite{lee2021technological} and zoology \cite{shen2021multivariate}, to name just a few.
 Especially in the field of unsupervised learning, clustering is utilized to extract structures from unlabeled data \cite{chander2023data}.
 
-Alongside the diverse applications of clustering, unique domain-specific requirements and challenges arise, each of which can be addressed by leveraging different families of clustering approaches \cite{xu2015comprehensive,han2012data}.
-These approaches, ranging from partitioning and hierarchical methods to density-based, grid-based and graph-based methods, are designed to accommodate varying data characteristics and application-related requirements to facilitate efficient cluster analyses across diverse domains.
+Alongside the diverse applications of clustering, unique domain-specific requirements and challenges arise, which are addressed by leveraging different families of clustering approaches \cite{xu2015comprehensive,han2012data}.
+These approaches, ranging from partitioning and hierarchical methods to density-based, grid-based and graph-based methods, are designed to accommodate varying data characteristics and domain-related requirements to facilitate efficient cluster analyses across diverse applications.
 
@@
 With the proliferation of complex data sources and the growing demand for more efficient clustering techniques, numerous \emph{fast k-means algorithms} have been introduced \cite{DBLP:conf/icml/Elkan03,DBLP:conf/sdm/Hamerly10,drake2012accelerated,hamerly2015accelerating,DBLP:conf/icml/NewlingF16,DBLP:conf/icml/DingZSMM15,DBLP:conf/sisap/SchubertLF21,DBLP:conf/sisap/LangS23}.
-These methods are designed to achieve exact clustering results with significantly reduced computational cost.
+These methods are designed to achieve good clustering results with significantly reduced computational cost.
 Moreover, most of them do not require any form of precomputation, making them directly applicable across a wide range of scenarios.
-The major objective of fast k-means algorithms is to reduce the number of distance evaluations required when assigning data objects to cluster centers and to safely exclude cluster centers from the assignment process.
+The major objective of fast k-means algorithms is to reduce the number of distance evaluations required when assigning data objects to cluster centers and to safely prune cluster centers from the assignment process.
 For this purpose, distances are approximated via lower and upper bounds.
@@
 With this extension, we do not aim to compete with the state of the art in k-means clustering;
-rather we intend to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations.
-We believe that our research findings offer promising directions for future research and new research impulses for data scientists and practitioners alike.
+rather, we intend to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations.
+We believe that our research findings offer a promising direction for future research and are beneficial for data scientists researchers and practitioners alike.
 
-This paper is structured as follows:
-Section~\ref{Preliminaries} outlines the k-means problem and Elkan's approach of acceleration the algorithmic computation.
-In Section~\ref{sec:contrib}, we proposes how to apply Ptolemy's inequality to the aforementioned algorithm as our main contribution.
+This short paper is structured as follows:
+Section~\ref{Preliminaries} outlines the k-means problem and Elkan's approach of accelerating the algorithmic computation.
+In Section~\ref{sec:contrib}, we show how Ptolemy's inequality can be applied to the aforementioned algorithm. %as our main contribution.
 The results of our preliminary performance evaluation are detailed in Section~\ref{sec: results}, while Section~\ref{sec: conclusions} concludes this paper with a short outlook on future work.
@@
 it would be valuable to explore adaptations of this technique for other k-means algorithms and optimizations.
-In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality can potentially be leveraged to achieve an enhancement in performance, making these clustering algorithms more practical for broader range of applications.
+In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality can potentially be leveraged to achieve an enhancement in performance, making these clustering algorithms more practical for a broader range of applications.
 

### 2024-11-15 review/rebuttal.md
add rebuttal draft and wait for feedback from NA,CB

Thank you for your reviews! Let me address your questions:

Total Runtime Measurements (R2, R3)
Our paper only considers the number of explicit distance evaluations to measure the efficiency of the algorithms we compare. Reviews 2 and 3 bring up the valid point of missing total execution time measurements. While we mention this fact as a limitation (in ch. 4.3, Limitations, point ii), our paper does not give a good explanation for this choice.

We chose to count distance evaluations because they are agnostic to the specific implementation of algorithms in machine code. The main disadvantage of reporting execution times is their strong dependence on the amount of optimization that went into the implementation.
For example, time measurements are affected by the programming language chosen, the amount of time spend on optimizing the source code, and the specific choice of hardware architecture (e.g. GPU or CPU).
Therefore, it takes considerable effort and care to make time measurements objective, reproducible, and generalizable. Consequently, in our short paper we focus on the theoretical, implementation-independent aspects of the discussed algorithms.

That said, the actual time measurements are important for real-world applications: Besides the distance evaluations, the caching and evaluation of the upper and lower bound approximations also takes time. While we reduce the total number of distance computations compared to Elkan's algorithm, our new bounds require more memory accesses and computations (cf. Eq. 3,4 and Eq. 5,6).
While this additional time is likely to be small compared to the time spent on distance evaluations, we plan to measure and report the actual ratio in a future full paper.

For the camera-ready version of our paper, we will include this elaboration and explicitly add execution time measurements as future work.

Discussion of Observed Behavior.
Review 3 notes that our discussion does not give a clear indication of how the chosen parameters relate to the speedups achieved, i.e. how the number of clusters $k$, the dataset, and its dimensionality influence the number of distance calculations saved. We agree that more precise rules predicting the actual speedups would be immensely useful.

However, we believe that we cannot, in good conscience, derive better rules from our data than those presented in the paper: While we did not cherry-pick the datasets used in our investigation, we support Review 3's notion that the speedups appear to be dataset-dependent (ch. 4.3, Limitations, point i). Similarly, we believe that a wider range of starting parameters (initializations, $k$s) is needed to draw more robust conclusions (ch. 4.3, Limitations, point iii).

Consequently, we feel confident that our discussion cannot go into more detail without running the significant risk of arriving at non-generalizable conclusions.

Analysis of the Set of Distance Computations.
Review 3 suggests an analysis of which distances are computed in each algorithm. We think this is an idea worth pursuing in the preparation of a full paper, but the actual implementation of this approach seems to require considerable effort.
While obtaining the data needed for such an analysis is straightforward, how to interpret it is still unclear to us at this time. Clusterings on data beyond the two-dimensional case are difficult to interpret visually, making it harder to develop meaningful measures for comparing the distance computation sets.

Minor Problems.
Review 3 correctly notes that Eq. 2 has a typo in the first summand of the RHS. We will correct this mistake in the camera-ready version. We will also improve the readability and accessibility of Fig. 1 by increasing its font size and using different line styles rather than relying on coloring alone.

Review 1 does not contain any questions or statements that need to be rebutted.

### 2024-11-17 review/rebuttal.md
second pass on rebuttal, incorporate CB criticism

@@
-Thank you for your reviews! Let me address your questions:
+Thank you for your reviews! We greatly appreciate your suggestions and will take them into consideration when updating our paper. Let me address your questions:
 
@@
 Reviews 2 and 3 bring up the valid point of missing total execution time measurements.
-While we mention this fact as a limitation (in ch.
+While we mention this fact as a limitation (ch.
 4.3, Limitations, point ii), our paper does not give a good explanation for this choice.
@@
 We chose to count distance evaluations because they are agnostic to the specific implementation of algorithms in machine code.
-The main disadvantage of reporting execution times is their strong dependence on the amount of optimization that went into the implementation.
+The main disadvantage of reporting execution times is their strong dependence on the amount of optimization that went into this implementation.
 For example, time measurements are affected by the programming language chosen, the amount of time spend on optimizing the source code, and the specific choice of hardware architecture (e.g.
 GPU or CPU).
-Therefore, it takes considerable effort and care to make time measurements objective, reproducible, and generalizable.
+It takes considerable effort and care to make time measurements objective, reproducible, and comparable.
 Consequently, in our short paper we focus on the theoretical, implementation-independent aspects of the discussed algorithms.
@@
 
-For the camera-ready version of our paper, we will include this elaboration and explicitly add execution time measurements as future work.
+Should the paper be accepted, we will include this elaboration and explicitly add execution time measurements as future work.
 
@@
 Review 3 correctly notes that Eq.
-2 has a typo in the first summand of the RHS.
-We will correct this mistake in the camera-ready version.
-We will also improve the readability and accessibility of Fig.
+2 has a typo in the first summand of the RHS, which will be corrected.
+Likewise, we will also improve the readability and accessibility of Fig.
 1 by increasing its font size and using different line styles rather than relying on coloring alone.

### 2025-01-08 review/rebuttal.md
rebuttal todo

@@
 
-That said, the actual time measurements are important for real-world applications: Besides the distance evaluations, the caching and evaluation of the upper and lower bound approximations also takes time.
+That said, the actual time measurements are important for real-world applications: Besides the distance evaluations, the caching, and evaluation of the upper and lower bound approximations also takes time.
 While we reduce the total number of distance computations compared to Elkan's algorithm, our new bounds require more memory accesses and computations (cf.

### 2025-01-08 paper.tex
add Makefile and fix references breaking the entire file
elaborate why we're not using time measurements

--- 2024-11-14 main.tex
@@
 \input{content/5_results}
-
-\section{Conclusions and Future Work} \label{sec: conclusions}
-
-% This paper addresses efficient data clustering using the $k$-means algorithm.
-% We extend Elkan's algorithm with novel distance approximations derived from Ptolemy's inequality.
-% Preliminary evaluations show significant performance improvements over Elkan's method, particularly with large numbers of clusters at low-to-medium dimensionalities.
-% The application of Ptolemy's inequality opens new avenues for future work, potentially enhancing other k-means algorithms that use triangle inequality for distance pruning.
-This approach could make clustering algorithms more practical across a broader range of applications.
-
-This short paper has addressed the problem of efficient data clustering by means of the k-means algorithm.
-We have shown how to extend the more efficient clustering algorithm proposed by Elkan with novel distance approximations derived from Ptolemy's inequality.
-The results of our preliminary performance evaluation indicate that our proposal achieves significant improvements in performance compared to Elkan's algorithm.
-%Particularly noteworthy is the increased efficiency with a large number of clusters.
-
-Moreover, the use of the Ptolemy's inequality opens up new avenues for future work.
-Besides conducting more extensive experiments to verify our findings,
-it would be valuable to explore adaptations of this technique for other k-means algorithms and optimizations.
-In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality can potentially be leveraged to achieve an enhancement in performance, making these clustering algorithms more practical for a broader range of applications.
-
-% \todo{Moreover, the use of Ptolemy’s inequality opens new avenues for future work.
-Exploring adaptations of this technique for other K-Means algorithms and optimizations could be highly valuable.
-In particular, algorithms that rely on the triangle inequality to prune distance calculations might leverage Ptolemy’s inequality to enhance performance, making these clustering algorithms more practical for a broader range of applications.}
-
+\input{content/6_conclusion.tex}
 \bibliography{references,bibliography_thesis,bibliography_iiwas}

### 2025-01-12 means.py
Update means.py
included reference to original code

@@ -32,7 +32,9 @@ class DataGenerator:
         labels = labels[permutation]
 
         return data, labels
-
+# This class mainly the elkan algorithm is derived and largely adapted from jjcordano's project (MIT License)
+# Original source: https://github.com/jjcordano/https://github.com/jjcordano/elkans_kmeans
+# new additions are the ptolemy algorithms
 class Kmeans:
     def __init__(self, k=2, max_iter=250, tolerance=0.000001, method='Elkan'):
         assert method in ['classic', 'Elkan', 'Ptolemy_upper', 'Ptolemy_lower', 'Ptolemy'], "Method argument not valid"

### 2025-01-12 main.py
added new main.py

@@ -1,0 +1,130 @@
+import numpy as np
+import pandas as pd
+from means import Kmeans, DataGenerator
+from sklearn.datasets import make_moons, make_circles, load_iris, load_wine
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+
+def generate_datasets():
+    datasets = {}
+
+    # Gaussian datasets (Blobs)
+    low_dim_generator = DataGenerator(n_samples=10000, n_features=10, n_clusters=4, random_state=43)
+    low_dim_data, _ = low_dim_generator.generate_data()
+    datasets["Gaussian_low"] = low_dim_data
+
+    medium_dim_generator = DataGenerator(n_samples=10000, n_features=100, n_clusters=4, random_state=43)
+    medium_dim_data, _ = medium_dim_generator.generate_data()
+    datasets["Gaussian_medium"] = medium_dim_data
+
+    high_dim_generator = DataGenerator(n_samples=10000, n_features=300, n_clusters=4, random_state=43)
+    high_dim_data, _ = high_dim_generator.generate_data()
+    datasets["Gaussian_high"] = high_dim_data
+
+    # Iris dataset
+    iris = load_iris()
+    datasets['Iris'] = iris.data
+
+    # Wine dataset
+    wine = load_wine()
+    datasets['Wine'] = wine.data
+
+    # Moons dataset
+    X_moons, _ = make_moons(n_samples=10000, noise=0.1, random_state=42)
+    datasets['Moons'] = X_moons
+
+    # Circles dataset
+    X_circles, _ = make_circles(n_samples=15000, noise=0.1, factor=0.5, random_state=42)
+    datasets['Circles'] = X_circles
+
+    # Random datasets
+    datasets["Random_low"] = np.random.uniform(0, 10, size=(5000, 10))
+    datasets["Random_medium"] = np.random.uniform(0, 10, size=(5000, 100))
+    datasets["Random_high"] = np.random.uniform(0, 10, size=(5000, 300))
+
+    return datasets
+
+
+def run_clustering(datasets, methods, k_values):
+    results = {"Dataset": [], "Method": [], "K": [], "Distance_Evaluations": []}
+    for name, data in datasets.items():
+        for k in k_values:
+            for method in methods:
+                model = Kmeans(k=k, method=method)
+                model.fit(data)
+                results["Dataset"].append(name)
[…]

### 2025-01-12 main.py
Adjusted structure

@@ -113,7 +113,7 @@ def plot_performance():
         ax.set_ylabel('Speedup')
 
     plt.tight_layout(rect=[0, 0, 0.9, 1])
-    plt.savefig('combined_plot_with_bigger_dataset_names.pdf', format='pdf')
+    plt.savefig('combined_plot.pdf', format='pdf')
     plt.show()
 
 

### 2025-01-13 main.py
add progress bar for main calculation

@@ -5,6 +5,7 @@ from sklearn.datasets import make_moons, make_circles, load_iris, load_wine
 import matplotlib.pyplot as plt
 import seaborn as sns
 
+from tqdm import tqdm
 
 def generate_datasets():
     datasets = {}
@@ -48,15 +49,19 @@ def generate_datasets():
 
 def run_clustering(datasets, methods, k_values):
     results = {"Dataset": [], "Method": [], "K": [], "Distance_Evaluations": []}
+    progress_bar = tqdm(total=len(datasets) * len(k_values) * len(methods))
     for name, data in datasets.items():
         for k in k_values:
             for method in methods:
+                progress_bar.set_postfix_str(f"{name}, k={k}, {method}")
                 model = Kmeans(k=k, method=method)
                 model.fit(data)
                 results["Dataset"].append(name)
                 results["Method"].append(method)
                 results["K"].append(k)
                 results["Distance_Evaluations"].append(model.distance_evaluations)
+
+                progress_bar.update()
     return pd.DataFrame(results)
 
 

### 2025-01-13 main.py
fix: save_results doesn't produce the two sheets that plot_performance requires

@@ -65,8 +65,14 @@ def run_clustering(datasets, methods, k_values):
     return pd.DataFrame(results)
 
 
-def save_results(results, filename="results/results.xlsx"):
-    results.to_excel(filename, index=False)
+def save_results(
+    results, filename="results/clustering_performance_results_formatted.xlsx"
+):
+    dim_suffixes = ("_low", "_medium", "_high")
+    has_dims = results["Dataset"].str.endswith(dim_suffixes)
+    with pd.ExcelWriter(filename, engine="openpyxl") as writer:
+        results[~has_dims].to_excel(writer, sheet_name="Results1", index=False)
+        results[has_dims].to_excel(writer, sheet_name="Results2", index=False)
 
 
 def plot_performance():

### 2025-01-13 main.py
also save wall clock times

@@ -5,6 +5,7 @@ from sklearn.datasets import make_moons, make_circles, load_iris, load_wine
 import matplotlib.pyplot as plt
 import seaborn as sns
 
+import time
 from tqdm import tqdm
 
 def generate_datasets():
@@ -48,18 +49,28 @@ def generate_datasets():
 
 
 def run_clustering(datasets, methods, k_values):
-    results = {"Dataset": [], "Method": [], "K": [], "Distance_Evaluations": []}
+    results = {
+        "Dataset": [],
+        "Method": [],
+        "K": [],
+        "Distance_Evaluations": [],
+        "Wall_Clock_Time": [],
+    }
     progress_bar = tqdm(total=len(datasets) * len(k_values) * len(methods))
     for name, data in datasets.items():
         for k in k_values:
             for method in methods:
                 progress_bar.set_postfix_str(f"{name}, k={k}, {method}")
+
                 model = Kmeans(k=k, method=method)
+                start_time = time.perf_counter()
                 model.fit(data)
+                end_time = time.perf_counter()
                 results["Dataset"].append(name)
                 results["Method"].append(method)
                 results["K"].append(k)
                 results["Distance_Evaluations"].append(model.distance_evaluations)
+                results["Wall_Clock_Time"].append(end_time - start_time)
 
                 progress_bar.update()
     return pd.DataFrame(results)

### 2025-01-13 plot_results.py
convert notebook code to script, call it from main.py

@@ -1,0 +1,120 @@
+import pandas as pd
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+
+def plot(file_path="results/clustering_performance_results_formatted.xlsx"):
+    data = pd.ExcelFile(file_path)
+    df = pd.concat(
+        [
+            data.parse("Results1"),
+            data.parse("Results2"),
+        ]
+    )
+
+    df.Dataset = df.Dataset.str.replace("Random", "Uniform")
+    df = df[df["K"].isin([3, 20, 100])]
+    df = df[~df.Dataset.isin(["Circles", "Moons"])]
+
+    # Function to calculate speedup
+    def calculate_speedup(group):
+        reference = group[group["Method"] == "Elkan"]["Distance_Evaluations"].values[0]
+        group["Speedup"] = reference / group["Distance_Evaluations"]
+        return group
+
+    all_speedups = (
+        df.groupby(["Dataset", "K"]).apply(calculate_speedup).reset_index(drop=True)
+    )
+    speedups = all_speedups[
+        all_speedups.Method.isin(["Ptolemy_upper", "Ptolemy_lower", "Ptolemy"])
+    ]
+
+    fontscale = 1
+    plt.rcParams.update(
+        {
+            "text.usetex": False,
+            "font.family": "sans-serif",
+            "font.size": 10 * fontscale,
+            "axes.titlesize": 10 * fontscale,
+            "axes.labelsize": 10 * fontscale,
+            "legend.fontsize": 10 * fontscale,
+            "xtick.labelsize": 10 * fontscale,
+            "ytick.labelsize": 10 * fontscale,
+        }
+    )
+
+    sns.set_palette("colorblind")
+
+    # width of available page is ~5 inches, we don't have a lot of height left, so only allocating 3 inches for height
+    scale = 1.7
+    fig = plt.figure(figsize=(5 * scale, 2.5 * scale), dpi=300, layout="constrained")
+    # fig = plt.figure(layout="constrained")
+
+    layout = [
+        ["Iris", "Wine", "legend"],
+        ["Gaussian_low", "Gaussian_medium", "Gaussian_high"],
+        ["Uniform_low", "Uniform_medium", "Uniform_high"],
+    ]
+    ax_mosaic = fig.subplot_mosaic(
+        layout,
+        sharex=True,
+    )
+
+    # sharey does not work as intended, so we have to set lables manually
+    for col in layout:
+        first = col[0]
+        for entry in col[1:]:
[…]

### 2025-01-14 paper/content/5_results.tex
put paper into its own subdir

--- 2024-11-14 content/5_results.tex
@@
 \subsection{Experiments}
-\todo{If we claim to have implemented the classic k-means, we also have to show that in our plots.}
 Our variant of the k-Means algorithm is tested by implementing the classical k-Means algorithm (Lloyd's algorithm), Elkan's algorithm, and its novel, Ptolemy-based extension in Python.
+
+Additionally, two intermediate variants were implemented to assess the separate effects of the lower and upper bounds:
+The \emph{Ptolemy Upper Bound} method uses Elkan's lower bound and Ptolemy's upper bound, while \emph{Ptolemy Lower Bound} uses Ptolemy's lower bound and Elkan's upper bound.
 
@@
 we measure the number of distance calculations instead of wall-clock time.
-As distance calculations account for the majority of computing time in k-Means algorithms, they offer a clear measure of efficiency regardless of implementational optimizations.
-
-Additionally, two intermediate variants were implemented to assess the separate effects of the lower and upper bounds:
-The \emph{Ptolemy Upper Bound} method uses Elkan's lower bound and Ptolemy's upper bound, while \emph{Ptolemy Lower Bound} uses Ptolemy's lower bound and Elkan's upper bound.
+Distance calculations account for the majority of computing time in k-Means algorithms and do not depend on the implementation of the algorithms to machine code.
+In comparison, execution time measurements are contingent on a number of interacting factors that render them difficult to reproduce.
+These include the programming language and version, the amount and distribution of time spend on optimizing each algorithm's implementation, and the hardware architecture used to conduct the measurements.
+Thus, it takes considerable effort and care to make time measurements objective, reproducible, and comparable.
+Consequently, this short paper focuses on the theoretical, implementation-independent aspects of the discussed algorithms to determine whether a more extensive study is justified.
 
@@
 The presented experiments are subject to three primary constraints:
-(i) While synthetic random datasets provide valuable best-case scenarios, more extensive real-world data with appropriate grouping structures is necessary for comprehensive evaluation.
+(i) While synthetic random datasets provide valuable best-case scenarios, more extensive real-world data with appropriate grouping structures is necessary for a comprehensive evaluation.
 (ii) The significant reduction in distance calculations may not directly translate to improved execution times, as Ptolemy's bounds are computationally more expensive than the triangle inequality.

### 2025-01-14 paper/content/6_conclusion.tex
put paper into its own subdir

--- 2024-10-02 main.tex
@@
-% !TeX encoding = UTF-8
-% !TeX program = pdflatex
-% !BIB program = bibtex
+\section{Conclusions and Future Work} \label{sec: conclusions}
+This short paper addresses the problem of efficient data clustering using the k-means algorithm.
+We show how the more efficient clustering algorithm proposed by Elkan can be extended with novel distance approximations derived from Ptolemy's inequality.
+Our preliminary performance evaluation, which compares the number of total distance calculations needed to cluster a dataset, indicates that our extension achieves a significant improvement in performance compared to Elkan's algorithm.
 
-\input{content/preamble}
+We see two primary avenues for future work:
+Firstly, more extensive experiments are required to verify our findings.
+Specifically, execution times on a wider range of datasets must be measured to determine under which conditions the extension leads to real-world runtime savings.
+Secondly, exploring adaptations of this approach to other k-means algorithms and optimizations could prove valuable.
+In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality might be leveraged to achieve better performance, making these clustering algorithms more practical for a broader range of applications.
 
-\begin{document}
-%%% Mehrere Autoren werden durch \and voneinander getrennt.
-%%% Die Fußnote enthält die Adresse sowie eine E-Mail-Adresse.
-%%% Das optionale Argument (sofern angegeben) wird für die Kopfzeile verwendet.
-\title[Extending K-Means Clustering with Ptolemy’s Inequality]{Extending K-Means Clustering with Ptolemy’s Inequality
-\thanks{This work is partly funded by DFG grant No. 454630593: EPIX – Efficient Ptolemaic Indexing.} 
-}
-%% \subtitle{Untertitel / Subtitle} % if needed
-\author[1]{Max Pernklau}{max.pernklau@fernuni-hagen.de}{0009-0007-5520-4093}
-\author[1]{Nikita Averitchev}{naveritchev@gmail.com}{}
-\author[1]{Christian Beecks}{christian.beecks@fernuni-hagen.de}{0009-0000-9028-629X}
-% \author[1]{Firstname4 Lastname4}{firstname4.lastname4@affiliation1.org}{0000-0000-0000-0000}%
-\affil[1]{FernUniversität in Hagen\\Chair of Data Science\\ 58084 Hagen\\Germany}
+%Particularly noteworthy is the increased efficiency with a large number of clusters.
 
-\maketitle
-\begin{abstract}
-	Clustering is a fundamental data analytics operation in the field of unsupervised learning.
-Given a database of unknown structure, clustering aims to discover the inherent structure of the data objects according to similarity, such that similar data objects are grouped together, while dissimilar ones are separated in different groups or clusters.
-	In this study, we focus on partitioning methods, specifically the \emph{k-means clustering} approach, which minimizes the intra-cluster variance.
-	As the standard algorithmic approach to k-means clustering, namely the Lloyd algorithm, is neither efficient nor scalable, various adaptations and modifications have been developed, resulting in the family of fast k-means clustering algorithms.
-In this short paper, we extend the clustering algorithm proposed by Elkan with Ptolemy's inequality to prune superfluous distance calculations.
-It is not our intention with this paper to compete with the state-of-the-art algorithmic solutions for the k-means clustering problem; rather, we seek to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, particularly in terms of reducing computations associated with distance evaluations.
-
-	% neither tri → pto nor pto → tri is true
-	% which generalizes the triangle inequality,
-
-	%where the Lloyd algorithm is a prominent and de facto standard implementation.
-
-	%The resulting groups, which are denoted as clusters, then represent the clustering structure of the database. A
-
-\end{abstract}
-\begin{keywords}
-	k-means clustering \and Lloyd algorithm \and Ptolemy's inequality %Keyword1 \and Keyword2
-\end{keywords}
-%%% Beginn des Artikeltexts
-
-%\todo{Americanize spelling (thesis uses chiefly British terms)}
-
-\section{Introduction}
-
-%intro
-Clustering is a fundamental data-analytical operation in the domain of data science.
-The basic task is to divide a set of data objects into different groups or clusters, such that objects within the same cluster have sufficient similarities;
-meanwhile objects in different clusters should have significant differences.
-The collection of clusters reflects the inherent structure of the database and is denoted as clustering.
-
-Clustering can be found in various scientific or economic application domains \cite{ezugwu2022comprehensive,oyewole2023data, gan2020data}, ranging from archaeology \cite{troiano2024comparative} and finance \cite{cai2016clustering} to industry \cite{lee2021technological} and zoology \cite{shen2021multivariate}, to name just a few.
-Especially in the field of unsupervised learning, clustering is utilized to extract structures from unlabeled data \cite{chander2023data}.
-
-Alongside the diverse applications of clustering, unique domain-specific requirements and challenges arise, each of which can be addressed by leveraging different families of clustering approaches \cite{xu2015comprehensive,han2012data}.
-These approaches, ranging from partitioning and hierarchical methods to density-based, grid-based and graph-based methods, are designed to accommodate varying data characteristics and application-related requirements to facilitate efficient cluster analyses across diverse domains.
-
-%different algorithms/problems
-%kmeans/lloyd
-%fast algorithms
-%inefficient: distance calculations
-%our proposal: ptolemaic
-%our purpose
-
-In this short paper, we focus on partitioning methods due to their simplicity in terms of interpretability and implementability \cite{DBLP:conf/iiwas/BeecksBHLSD22}.
-In this field, the \emph{k-means algorithm} \cite{bock2007clustering,hans2008origins,DBLP:journals/prl/Jain10,steinley2006k} has become one of the most influential clustering techniques \cite{DBLP:journals/kais/WuKQGYMMNLYZSHS08,olukanmi2019rethinking}.
-Although the term \emph{k-means algorithm} is explicitly credited to MacQueen \cite{macqueen1967}, its origins trace back to Steinhaus \cite{steinhaus1956division} in 1956, with its first application to data clustering by Forgy \cite{forgy1965cluster} in 1965.
-The widely recognized version in use today is the \emph{Lloyd algorithm} \cite{DBLP:journals/tit/Lloyd82}, introduced in 1982.
-
-With the proliferation of complex data sources and the growing demand for more efficient clustering techniques, numerous \emph{fast k-means algorithms} have been introduced \cite{DBLP:conf/icml/Elkan03,DBLP:conf/sdm/Hamerly10,drake2012accelerated,hamerly2015accelerating,DBLP:conf/icml/NewlingF16,DBLP:conf/icml/DingZSMM15,DBLP:conf/sisap/SchubertLF21,DBLP:conf/sisap/LangS23}.
-These methods are designed to achieve exact clustering results with significantly reduced computational cost.
-Moreover, most of them do not require any form of precomputation, making them directly applicable across a wide range of scenarios.
-The major objective of fast k-means algorithms is to reduce the number of distance evaluations required when assigning data objects to cluster centers and to safely exclude cluster centers from the assignment process.
-For this purpose, distances are approximated via lower and upper bounds.
-
-The Elkan algorithm \cite{DBLP:conf/icml/Elkan03} is a prominent representative of these algorithms and particularly suited for high-dimensional scenarios.
-It maintains one upper bound and multiple lower bounds for each data object in order to apply different pruning criteria for each combination of data object and cluster center.
-While the lower and upper bounds of Elkan's algorithm are directly derived from the triangle inequality, we propose to utilitze Ptolemy's inequality instead in order to increase the pruning performance.
-With this extension, we do not aim to compete with the state of the art in k-means clustering;
-rather we intend to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations.
-We believe that our research findings offer promising directions for future research and new research impulses for data scientists and practitioners alike.
-
-This paper is structured as follows:
-Section~\ref{Preliminaries} outlines the k-means problem and Elkan's approach of acceleration the algorithmic computation.
-In Section~\ref{sec:contrib}, we proposes how to apply Ptolemy's inequality to the aforementioned algorithm as our main contribution.
-The results of our preliminary performance evaluation are detailed in Section~\ref{sec: results}, while Section~\ref{sec: conclusions} concludes this paper with a short outlook on future work.
-
-%contribution
-
-%structure
-
-%\newpage
-
-%\section{Related Work}
-%\todo{remove this section if we run out of time}
-
-\input{content/3_preliminaries}
-\input{content/4_contribution}
-
-\section{Preliminary Results}\label{sec: results}
-
-\input{content/5_results}
-
-\section{Conclusions and Future Work} \label{sec: conclusions}
+% \todo{Moreover, the use of Ptolemy’s inequality opens new avenues for future work.
+Exploring adaptations of this technique for other K-Means algorithms and optimizations could be highly valuable.
+In particular, algorithms that rely on the triangle inequality to prune distance calculations might leverage Ptolemy’s inequality to enhance performance, making these clustering algorithms more practical for a broader range of applications.}
 
@@
 This approach could make clustering algorithms more practical across a broader range of applications.
-
-This short paper has addressed the problem of efficient data clustering by means of the k-means algorithm.
-We have shown how to extend the more efficient clustering algorithm proposed by Elkan with novel distance approximations derived from Ptolemy's inequality.
-The results of our preliminary performance evaluation indicate that our proposal achieves significant improvements in performance compared to Elkan's algorithm.
-%Particularly noteworthy is the increased efficiency with a large number of clusters.
-
-Moreover, the use of the Ptolemy's inequality opens up new avenues for future work.
-Besides conducting more extensive experiments to verify our findings,
-it would be valuable to explore adaptations of this technique for other k-means algorithms and optimizations.
-In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality can potentially be leveraged to achieve an enhancement in performance, making these clustering algorithms more practical for broader range of applications.
-
-% \todo{Moreover, the use of Ptolemy’s inequality opens new avenues for future work.
-Exploring adaptations of this technique for other K-Means algorithms and optimizations could be highly valuable.
-In particular, algorithms that rely on the triangle inequality to prune distance calculations might leverage Ptolemy’s inequality to enhance performance, making these clustering algorithms more practical for a broader range of applications.}
-
-\bibliography{references,bibliography_thesis,bibliography_iiwas}
-\end{document}

### 2025-01-14 paper/content/alg_elkan.tex
put paper into its own subdir

--- 2024-10-01 content/3u4_methods.tex
@@
-\section{Foundations of Ptolemaic Indexing}
-
-\todo{explain Hetland's lower (upper) bound notation ($d^-$ and $d^+$)}
-\begin{theorem}[Ptolemaic Bounds]
-	\label{thm4.2}
-	\todo{we might want to explain what the points are supposed to do before showing this theorem.
-		Or, alternatively, we just present this as a purely mathematical statement with no preceding motivation.}
-	Let \(a, b, c, c' \in \mathbb{R}^n\) and let \(d(\cdot, \cdot)\) be the Euclidean distance. The distance \(d(a, c')\) can be bounded as follows:
-	\begin{gather}
-		\frac{1}{d(b, c)} \cdot \max \left\{
-		\begin{array}{l}
-			d(a, b)^- \cdot d(c, c') - d(a, c)^+ \cdot d(b, c'), \\
-			d(a, c)^- \cdot d(b, c') - d(a, b)^+ \cdot d(c, c')
-		\end{array}
-		\right\} \leq d(a, c')
-	\end{gather}
-	and
-	\begin{gather}
-		d(a, c') \leq \frac{1}{d(b, c)} \cdot \left( d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c') \right)
-	\end{gather}
-\end{theorem}
-\begin{proof}
-	\todo{convert to passive voice}
-	\todo{expand context to summarize all steps that lead to this proof}
-	The lower bound corresponds to the derivation in equation \ref{eq4.4} and corresponds to the arrangement of the two pivot objects.
-Replace some distances with the respective lower or upper bound for the distance.
-The lower bounds can be used as they are always smaller than the exact distance.
-The upper bound must be used due to the subtraction.
-This means that the inequality still applies.
-The upper bound is obtained by using Ptolemy's inequality:
-
-	\begin{equation*}
-		\begin{aligned}
-			\label{eq4.7}
-			d(a,c') \cdot d(b,c) & \leq d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')                \\
-			d(a,c')              & \leq \frac{d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')}{d(b,c)}
-		\end{aligned}
-	\end{equation*}
-	By replacing some distances with upper boundaries you get the desired upper boundaries.
-\end{proof}
-
-\section{Applying Ptolemy's Inequality to the Elkan's Algorithm}
-
-\subsection{Integration in k-Means}
-Theorem \ref{thm4.2} will now be applied in a similar context to Elkan's algorithm \ref{alg:elkan}, with the goal of integrating these new bounds into steps 5 and 6 of the algorithm.
-
-Consider the data set \(D = \{x_1, \ldots, x_n\}\).
-For each \(x_i\), there is an upper \(u_i\) and lower bounds \(l_{ij}\) as in Elkan's algorithm.
-The new cluster center assigned in a iteration is denoted by \(c_i^{\text{new}}\), and the old cluster center by \(c_i^{\text{old}}\).
-To use Ptolemy's Inequality, one additional point is needed.
-The idea is to use an even older cluster center \(c_i^{\text{old,old}}\) obtained two iterations ago.
-
-Overall, to obtain a new upper bound for the distance between the datapoint \(x_i\) and the new cluster center \(c_j^{\text{new}}\), consider the old cluster center \(c_j^{\text{old}}\) from the last iteration to which \(x_i\) was assigned, and the even older cluster center \(c_j^{\text{old,old}}\) from two iterations ago.
-Utilize the old upper bounds that were valid for the respective iteration.
-
-\begin{equation*}
-	\begin{aligned}
-		d(x_i,c_i^{old})      & \leq u_i^{old}      \\
-		d(x_i,c_i^{old, old}) & \leq u_i^{old, old}
-	\end{aligned}
-\end{equation*}
-
-Similarly, to obtain a new lower bound for the distance between the datapoint \(x_i\) and the new cluster center \(c_j^{\text{new}}\), the old lower bounds from the respective iterations are used:
-
-\begin{equation*}
-	\begin{aligned}
-		l_{i,j}^{\text{old}}     & \leq d(x_i, c_j^{\text{old}})     \\
-		l_{i,j}^{\text{old,old}} & \leq d(x_i, c_j^{\text{old,old}})
-	\end{aligned}
-\end{equation*}
-
-Using Theorem \ref{thm4.2}, the following upper bound for the distance are obtained:
-\begin{equation}
-	\label{eq4.7}
-	d(x_i, c_j^{\text{new}}) \leq u_i^{\text{new}} = \frac{1}{d(c_j^{\text{old}}, c_j^{\text{old,old}})} \cdot \left( u_i^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}) + u_i^{\text{old,old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old}}) \right)
-\end{equation}
-
-and the new lower bound:
-\begin{equation}
-	\label{eq4.8}
-	d(x_i, c_j^{\text{new}}) \geq l_{i,j}^{\text{new}} = \frac{1}{d(c_j^{\text{old}}, c_j^{\text{old,old}})} \cdot \max \left\{
-	\begin{array}{l}
-		l_{i,j}^{\text{old,old}} \cdot d(c_j^{\text{old}}, c_j^{\text{new}}) - u_i^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}), \\
-		l_{i,j}^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}) - u_i^{\text{old,old}} \cdot d(c_j^{\text{old}}, c_j^{\text{new}})
-	\end{array}
-	\right\}
-\end{equation}
-
-You can see that the old upper boundaries are required to calculate the new lower boundaries.
-
-Let us now look at how these bounds can be integrated into Elkan's algorithm.
-It is important to save not only the cluster centres and the bounds of the last iteration, but also those of the penultimate iteration.
-
-At the beginning, all distances to all cluster centres are calculated in the first iteration.
-The upper bound is set by the distance to the nearest cluster centre, and the lower bounds are determined by the direct distances from the point to the respective cluster centres.
-An iteration is then performed as with Elkan, which means that bounds from two iterations are now available.
-This makes it possible to apply the new bounds from equations \ref{eq4.7} and \ref{eq4.8}.
-The algorithm then iterates, as in the original Elkan algorithm using updating with new update conditions, until the cluster centres converge or a specified number of iterations is reached.
-It is always important to check whether $d(c_j^{\text{old}}, c_j^{\text{old,old}}) \neq 0$.
-If this is the case, the update is carried out as for Elkan.
-The algorithm is displayed in Algorithm \ref{alg:ptolemy}.
-
+\todo{original Elkan as presented in NiAv's bachelor thesis. has slightly excessive spacing}
 \begin{algorithm}
-	\caption{Elkan's Algorithm extended with Ptolemy's inequality}
-	\label{alg:ptolemy}
-
+	\caption{Accelerated k-Means Algorithm (Elkan)}
+	\label{alg:elkan}
 	\begin{algorithmic}
@@
 
-		\State \textbf{Step 0:} Compute one iteration of Elkan's algorithm. Save old upper and lower bounds of the last two iterations.
 		\Repeat
@@
 		\Statex
-		\State \textbf{Step 5:} If distance between two older cluster is not 0.
-Update lower bounds by using equation \ref{eq4.8}.
-Otherwise, update like in Elkan.
-Save the old lower bounds of the two last iterations.
+		\State \textbf{Step 5:} Update lower bounds $l(x_i, c_j)$ for each $x_i$ and $c_j$ using $l(x_i, c_j) = \max \{ l(x_i, c_j) - d(c'_j, c_j), 0 \}$.
 		\Statex
-		\State \textbf{Step 6:} If distance between two older cluster is not 0.
-Update upper bounds by using equation \ref{eq4.7}.
-Otherwise, update like in Elkan.
-Reset $r(x_i) = \text{true}$.
-Save old cluster and upper bounds of the last two iterations.
+		\State \textbf{Step 6:} Update upper bounds $u(x_i)$ for each $x_i$ using $u(x_i) = u(x_i) + d(c'(x_i), c(x_i))$ and reset $r(x_i) = \text{true}$.
 		\Statex

### 2025-01-14 paper/content/5_results.tex
fix: speedup defined the wrong way around
fix figure description

@@
 we report our results in the form of speedups,
-i.e. as the number of distances calculated by our extension,
-divided by the number of distances calculated by Elkan's algorithm.
+i.e. as the number of distances calculated by Elkan's algorithm,
+divided by the number of distances  calculated by our extension.
 
@@
         A speedup of $>1$ indicates an improvement over Elkan's original algorithm.
-		\textbf{Top and middle row:} Performance increases on synthetic data generated from Gaussian (top) and uniformly random (middle) distributions.
+		\textbf{Top row:} Performance increases on the real-world datasets Iris (left) and Wine (center).
+		\textbf{Center and bottom row:} Performance increases on synthetic data generated from Gaussian (center row) and uniformly random (bottom row) distributions.
 		The dimensionality of the dataset increases left-to-right.
-		\textbf{Bottom row:} Performance increases on the real-world datasets Iris (left) and Wine (right).
 	}

### 2025-01-14 paper/content/4_contribution.tex

--- 2024-10-02 content/4_contribution.tex
@@
 Elkan's algorithm can improve the execution speed of $k$-means clustering through the use of the triangle inequality.
-Explicit distance calculations are avoided when tight upper and lower bounds to the distance function are available.
+Explicit distance calculations are avoided when tight upper and lower bounds of the distance function are available.
 
 In this section, we use Ptolemy's inequality to calculate an additional set of bounds and modify Elkan's algorithm accordingly.
-These Ptolemaic bounds often improve upon the triangular bounds, leading to increased execution speeds, as will be shown in \autoref{sec:results}.
+These Ptolemaic bounds often improve upon the triangular bounds, leading to increased execution speeds, as will be shown in Section \ref{sec: results}.
 
@@
 while Ptolemy's inequality does so for four points.
-At the end of an iteration of Elkan's algorithm (Step~5~to~7 in \autoref{alg:elkan}),
-the distance bounds between the data points $x_i$ and the new center positions are calculated $c'_j$.
+At the end of an iteration of Elkan's algorithm (Step~5~to~7 in Alg.~\ref{alg:elkan}),
+the distance bounds between the data points $x_i$ and the new center positions $c'_j$ are calculated.
 Therefore, the three points used in the calculation of the bounds are the aforementioned two points and $c_j$, the cluster center calculated in the previous iteration.
@@
 	\end{align}
-	where $u_i$ and $u^\prev_i$ are the lower bounds calculated in the previous and the penultimate iteration, respectively.
+	where $u_i, u^\prev_i$ ($l_{i,j}, l^\prev_{i,j}$) are the lower (upper) bounds calculated in the previous and the penultimate iteration, respectively\footnote{
+ For readability, we suppress the bounds' arguments in favor of indices, e.g. $u_i$ instead of $u(x_i)$.
+    }.
 \end{theorem}
+\vspace{-2\baselineskip}
 \begin{proof}
-	As all points are in the $\mathrm{R^n}$ and the distances function is also Euclidean, Ptolemy's inequality
+	As all points are in the $\mathbb{R}^n$ and the distances function is also Euclidean, Ptolemy's inequality
 	$d(x, y)\cdot d(v, u) \leq d(x, v) \cdot d(y,v) + d(x, u) \cdot d(y, v)$ holds.
@@
 	The process is similar for Eq.~\ref{eq:pto_lower}; as the signs on the right-hand side are not equal in this case, there are two distinct ways to substitute into Ptolemy's inequality.
-	This gives rise to the two inequalities, of which the maximum is the stronger bound.
-	Analogous to $u'_i$, the right-hand side of Eq.~\ref{eq:pto_lower} can only be made smaller by inserting lower bounds of the distance $l_{i,j},l_{i,j}^\prev$ in the minuend or upper bounds $u_i, u^\prev_i$ in the subtrahend.
+	This gives rise to two inequalities, of which the maximum is the stronger bound.
+	Analogous to $u'_i$, the right-hand side of Eq.~\ref{eq:pto_lower} can only be made smaller by inserting lower bounds $l_{i,j},l_{i,j}^\prev$ in the minuend or upper bounds $u_i, u^\prev_i$ in the subtrahend.
 	Thus, $d(x_i, c_j') \geq l_{i,j}'$ also holds.
 \end{proof}
+
+\subsection{Integration}
+To integrate the novel bounds into the existing framework of Elkan's algorithm,
+ Steps~6 and~7 in Alg.~\autoref{alg:elkan} need to be updated to use the new bounds.
+It is possible to only replace either the upper or lower bound, which leads to a hybrid solution that is further explored in Section \ref{sec: results}.
+
+As the Ptolemaic bounds require information from two previous iterations, the first iteration of the algorithm is always conducted with Elkan's original bounds.
 

### 2025-01-14 paper/content/alg_pto.tex

--- 2024-10-01 content/3u4_methods.tex
@@
-\section{Foundations of Ptolemaic Indexing}
-
-\todo{explain Hetland's lower (upper) bound notation ($d^-$ and $d^+$)}
-\begin{theorem}[Ptolemaic Bounds]
-	\label{thm4.2}
-	\todo{we might want to explain what the points are supposed to do before showing this theorem.
-		Or, alternatively, we just present this as a purely mathematical statement with no preceding motivation.}
-	Let \(a, b, c, c' \in \mathbb{R}^n\) and let \(d(\cdot, \cdot)\) be the Euclidean distance. The distance \(d(a, c')\) can be bounded as follows:
-	\begin{gather}
-		\frac{1}{d(b, c)} \cdot \max \left\{
-		\begin{array}{l}
-			d(a, b)^- \cdot d(c, c') - d(a, c)^+ \cdot d(b, c'), \\
-			d(a, c)^- \cdot d(b, c') - d(a, b)^+ \cdot d(c, c')
-		\end{array}
-		\right\} \leq d(a, c')
-	\end{gather}
-	and
-	\begin{gather}
-		d(a, c') \leq \frac{1}{d(b, c)} \cdot \left( d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c') \right)
-	\end{gather}
-\end{theorem}
-\begin{proof}
-	\todo{convert to passive voice}
-	\todo{expand context to summarize all steps that lead to this proof}
-	The lower bound corresponds to the derivation in equation \ref{eq4.4} and corresponds to the arrangement of the two pivot objects.
-Replace some distances with the respective lower or upper bound for the distance.
-The lower bounds can be used as they are always smaller than the exact distance.
-The upper bound must be used due to the subtraction.
-This means that the inequality still applies.
-The upper bound is obtained by using Ptolemy's inequality:
-
-	\begin{equation*}
-		\begin{aligned}
-			\label{eq4.7}
-			d(a,c') \cdot d(b,c) & \leq d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')                \\
-			d(a,c')              & \leq \frac{d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')}{d(b,c)}
-		\end{aligned}
-	\end{equation*}
-	By replacing some distances with upper boundaries you get the desired upper boundaries.
-\end{proof}
-
-\section{Applying Ptolemy's Inequality to the Elkan's Algorithm}
-
-\subsection{Integration in k-Means}
-Theorem \ref{thm4.2} will now be applied in a similar context to Elkan's algorithm \ref{alg:elkan}, with the goal of integrating these new bounds into steps 5 and 6 of the algorithm.
-
-Consider the data set \(D = \{x_1, \ldots, x_n\}\).
-For each \(x_i\), there is an upper \(u_i\) and lower bounds \(l_{ij}\) as in Elkan's algorithm.
-The new cluster center assigned in a iteration is denoted by \(c_i^{\text{new}}\), and the old cluster center by \(c_i^{\text{old}}\).
-To use Ptolemy's Inequality, one additional point is needed.
-The idea is to use an even older cluster center \(c_i^{\text{old,old}}\) obtained two iterations ago.
-
-Overall, to obtain a new upper bound for the distance between the datapoint \(x_i\) and the new cluster center \(c_j^{\text{new}}\), consider the old cluster center \(c_j^{\text{old}}\) from the last iteration to which \(x_i\) was assigned, and the even older cluster center \(c_j^{\text{old,old}}\) from two iterations ago.
-Utilize the old upper bounds that were valid for the respective iteration.
-
-\begin{equation*}
-	\begin{aligned}
-		d(x_i,c_i^{old})      & \leq u_i^{old}      \\
-		d(x_i,c_i^{old, old}) & \leq u_i^{old, old}
-	\end{aligned}
-\end{equation*}
-
-Similarly, to obtain a new lower bound for the distance between the datapoint \(x_i\) and the new cluster center \(c_j^{\text{new}}\), the old lower bounds from the respective iterations are used:
-
-\begin{equation*}
-	\begin{aligned}
-		l_{i,j}^{\text{old}}     & \leq d(x_i, c_j^{\text{old}})     \\
-		l_{i,j}^{\text{old,old}} & \leq d(x_i, c_j^{\text{old,old}})
-	\end{aligned}
-\end{equation*}
-
-Using Theorem \ref{thm4.2}, the following upper bound for the distance are obtained:
-\begin{equation}
-	\label{eq4.7}
-	d(x_i, c_j^{\text{new}}) \leq u_i^{\text{new}} = \frac{1}{d(c_j^{\text{old}}, c_j^{\text{old,old}})} \cdot \left( u_i^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}) + u_i^{\text{old,old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old}}) \right)
-\end{equation}
-
-and the new lower bound:
-\begin{equation}
-	\label{eq4.8}
-	d(x_i, c_j^{\text{new}}) \geq l_{i,j}^{\text{new}} = \frac{1}{d(c_j^{\text{old}}, c_j^{\text{old,old}})} \cdot \max \left\{
-	\begin{array}{l}
-		l_{i,j}^{\text{old,old}} \cdot d(c_j^{\text{old}}, c_j^{\text{new}}) - u_i^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}), \\
-		l_{i,j}^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}) - u_i^{\text{old,old}} \cdot d(c_j^{\text{old}}, c_j^{\text{new}})
-	\end{array}
-	\right\}
-\end{equation}
-
-You can see that the old upper boundaries are required to calculate the new lower boundaries.
-
-Let us now look at how these bounds can be integrated into Elkan's algorithm.
-It is important to save not only the cluster centres and the bounds of the last iteration, but also those of the penultimate iteration.
-
-At the beginning, all distances to all cluster centres are calculated in the first iteration.
-The upper bound is set by the distance to the nearest cluster centre, and the lower bounds are determined by the direct distances from the point to the respective cluster centres.
-An iteration is then performed as with Elkan, which means that bounds from two iterations are now available.
-This makes it possible to apply the new bounds from equations \ref{eq4.7} and \ref{eq4.8}.
-The algorithm then iterates, as in the original Elkan algorithm using updating with new update conditions, until the cluster centres converge or a specified number of iterations is reached.
-It is always important to check whether $d(c_j^{\text{old}}, c_j^{\text{old,old}}) \neq 0$.
-If this is the case, the update is carried out as for Elkan.
-The algorithm is displayed in Algorithm \ref{alg:ptolemy}.
-
+\todo{elkan's alg. changed to inlcude Pto LBs. too large to include in the paper}
 \begin{algorithm}

### 2025-01-14 plot_results.py
plotting: fix warnings, add grid

@@ -3,8 +3,11 @@ import matplotlib.pyplot as plt
 import seaborn as sns
 
 
-def plot(file_path="results/clustering_performance_results_formatted.xlsx"):
-    data = pd.ExcelFile(file_path)
+def plot(
+    input_file="results/clustering_performance_results_formatted.xlsx",
+    output_file="results/combined_plot.pdf",
+):
+    data = pd.ExcelFile(input_file)
     df = pd.concat(
         [
             data.parse("Results1"),
@@ -17,13 +20,17 @@ def plot(file_path="results/clustering_performance_results_formatted.xlsx"):
     df = df[~df.Dataset.isin(["Circles", "Moons"])]
 
     # Function to calculate speedup
+    metric = "Distance_Evaluations"
+
     def calculate_speedup(group):
-        reference = group[group["Method"] == "Elkan"]["Distance_Evaluations"].values[0]
-        group["Speedup"] = reference / group["Distance_Evaluations"]
+        reference = group[group["Method"] == "Elkan"][metric].values[0]
+        group["Speedup"] = reference / group[metric]
         return group
 
     all_speedups = (
-        df.groupby(["Dataset", "K"]).apply(calculate_speedup).reset_index(drop=True)
+        df.groupby(["Dataset", "K"])[df.columns]
+        .apply(calculate_speedup)
+        .reset_index(drop=True)
     )
     speedups = all_speedups[
         all_speedups.Method.isin(["Ptolemy_upper", "Ptolemy_lower", "Ptolemy"])
@@ -91,6 +98,7 @@ def plot(file_path="results/clustering_performance_results_formatted.xlsx"):
             ax=axis,
             **plt_kwargs,
         )
+        axis.grid(visible=True, axis="y")
         axis.set_xlabel("$k$")
         axis.set_title(name.replace("_", ", "))
 
@@ -117,4 +125,9 @@ def plot(file_path="results/clustering_performance_results_formatted.xlsx"):
         yticks.sort()
         axis.set_yticks(yticks)
 
-    plt.savefig("results/combined_plot.pdf", format="pdf")
+    plt.savefig(output_file, format="pdf")
+
+
+if __name__ == "__main__":
+    print("Producing plot")
+    plot()

### 2025-01-14 paper/content/5_results.tex

@@
 we measure the number of distance calculations instead of wall-clock time.
-Distance calculations account for the majority of computing time in k-Means algorithms and do not depend on the implementation of the algorithms to machine code.
+Distance calculations account for the majority of computing time in k-Means algorithms and do not depend on the specific machine-code implementation.
 In comparison, execution time measurements are contingent on a number of interacting factors that render them difficult to reproduce.
-These include the programming language and version, the amount and distribution of time spend on optimizing each algorithm's implementation, and the hardware architecture used to conduct the measurements.
+These include the programming language and version, the amount and distribution of time spend on optimizing each algorithm's implementation, and the hardware used to conduct such measurements.
 Thus, it takes considerable effort and care to make time measurements objective, reproducible, and comparable.
-Consequently, this short paper focuses on the theoretical, implementation-independent aspects of the discussed algorithms to determine whether a more extensive study is justified.
+Consequently, this short paper focuses only on the theoretical, implementation-independent aspects of the discussed algorithms to determine whether a more extensive study is justified.
 
@@
 The full Ptolemaic algorithm consistently demonstrates significant improvements over Elkan’s algorithm as the number of clusters $k$ increases.
-For instance, in the Gaussian low-dimensional dataset, the Ptolemaic algorithm achieves larger than 2-fold speedups for $k = 20$ and $k = 100$, illustrating its ability to significantly reduce distance calculations in complex clustering tasks.
+For instance, in the Gaussian low-dimensional dataset, the Ptolemaic algorithm achieves at least a 2-fold speedup for $k = 20$ and $k = 100$, illustrating its ability to significantly reduce distance calculations in complex clustering tasks.
 % surpasses 3-fold improvements for
 % Similar trends are observed in the Random low-dimensional dataset.
-For smaller cluster counts ($k = 3$), the advantage is less pronounced, with speedups around 1.1 in the Iris and Wine datasets.
-This suggests that Ptolemy's inequality offers significant improvements for larger values of $k$.
-However, lower-complexity clustering tasks may not benefit to the same extent,
+For smaller cluster counts ($k = 3$), the advantage is less pronounced: We see a significant difference in performance only in the low-dimensional Guassian dataset, while all other datasets show no significant increases or decreases.
+This suggests that Ptolemy's inequality offers significant improvements only for larger values of $k$.
+Lower-complexity clustering tasks might not benefit to the same extent,
 given the additional overhead imposed by the Ptolemaic bounds, which are computationally more complex to evaluate compared to the triangle bounds.
-The Ptolemaic Upper Bound variant performs comparably to the full Ptolemaic algorithm, especially for larger $k$ values.
-However, the Lower Bound variant underperforms in some cases, particularly with smaller cluster counts or in higher-dimensional datasets.
+The Ptolemaic Upper Bound variant performs roughly similar to the full Ptolemaic algorithm.
+In comparison, the Lower Bound variant has comparable or worse performance to Elkan.
 
-\textbf{Dimensionality.} The full Ptolemaic algorithm performs well across varying dataset dimensionality.
+\textbf{Dimensionality.} The full Ptolemaic algorithm performs reasonably well across varying dataset dimensionality.
 In low-dimensional datasets (10 dimensions), it achieves impressive speedups, particularly at higher cluster counts.

### 2025-01-14 paper/content/6_conclusion.tex

@@
 Firstly, more extensive experiments are required to verify our findings.
-Specifically, execution times on a wider range of datasets must be measured to determine under which conditions the extension leads to real-world runtime savings.
+Specifically, execution times---instead of distance calculation counts---on a wider range of datasets must be measured to determine under which conditions the extension leads to real-world runtime savings.
 Secondly, exploring adaptations of this approach to other k-means algorithms and optimizations could prove valuable.

### 2025-01-15 paper/content/2_introduction.tex
split out introduction

--- 2025-01-08 paper.tex
@@
-% !TeX encoding = UTF-8
-% !TeX program = pdflatex
-% !BIB program = bibtex
-
-\input{content/preamble}
-
-\begin{document}
-%%% Mehrere Autoren werden durch \and voneinander getrennt.
-%%% Die Fußnote enthält die Adresse sowie eine E-Mail-Adresse.
-%%% Das optionale Argument (sofern angegeben) wird für die Kopfzeile verwendet.
-\title[Extending K-Means Clustering with Ptolemy’s Inequality]{Extending K-Means Clustering with Ptolemy’s Inequality
-	\thanks{This work is partly funded by DFG grant No. 454630593: EPIX – Efficient Ptolemaic Indexing.}
-}
-%% \subtitle{Untertitel / Subtitle} % if needed
-\author[1]{Max Pernklau}{max.pernklau@fernuni-hagen.de}{0009-0007-5520-4093}
-\author[1]{Nikita Averitchev}{naveritchev@gmail.com}{}
-\author[1]{Christian Beecks}{christian.beecks@fernuni-hagen.de}{0009-0000-9028-629X}
-% \author[1]{Firstname4 Lastname4}{firstname4.lastname4@affiliation1.org}{0000-0000-0000-0000}%
-\affil[1]{FernUniversität in Hagen\\Chair of Data Science\\ 58084 Hagen\\Germany}
-
-\maketitle
-\begin{abstract}
-	Clustering is a fundamental data analytics operation in the field of unsupervised learning.
-Given a database of unknown structure, clustering aims to discover the inherent structure of the data objects according to similarity, such that similar data objects are grouped together, while dissimilar ones are separated in different groups or clusters.
-	In this study, we focus on partitioning methods, specifically the \emph{k-means clustering} approach, which minimizes the intra-cluster variance.
-	As the standard algorithmic approach to k-means clustering, the Lloyd algorithm, is neither efficient nor scalable, various adaptations and modifications have been developed, resulting in the family of fast k-means clustering algorithms.\\
-	In this short paper, we extend the clustering algorithm proposed by Elkan with Ptolemy's inequality to prune superfluous distance calculations.
-It is not our intention to compete with the state-of-the-art algorithmic solutions for the k-means clustering problem; rather, we seek to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, particularly in terms of reducing computations associated with distance evaluations.
-
-	% neither tri → pto nor pto → tri is true
-	% which generalizes the triangle inequality,
-
-	%where the Lloyd algorithm is a prominent and de facto standard implementation.
-
-	%The resulting groups, which are denoted as clusters, then represent the clustering structure of the database. A
-
-\end{abstract}
-\begin{keywords}
-	k-means clustering \and Lloyd algorithm \and Ptolemy's inequality %Keyword1 \and Keyword2
-\end{keywords}
-%%% Beginn des Artikeltexts
-
-%\todo{Americanize spelling (thesis uses chiefly British terms)}
-
 \section{Introduction}
@@
 %\todo{remove this section if we run out of time}
-
-\input{content/3_preliminaries}
-\input{content/4_contribution}
-
-\section{Preliminary Results}\label{sec: results}
-
-\input{content/5_results}
-\input{content/6_conclusion.tex}
-\bibliography{references,bibliography_thesis,bibliography_iiwas}
-\end{document}

### 2025-01-15 paper/paper.tex
split out introduction

--- 2025-01-08 paper.tex
@@
 
-\section{Introduction}
-
-%intro
-Clustering is a fundamental data-analytics operation in the domain of data science.
-The basic task is to divide a set of data objects into different groups or clusters, such that objects within the same cluster have sufficient similarities;
-meanwhile objects in different clusters should have significant differences.
-The collection of clusters reflects the inherent structure of the underlying data-generating process and is denoted as a clustering.
-
-The need for clustering arises in various scientific or economic application domains \cite{ezugwu2022comprehensive,oyewole2023data, gan2020data}, ranging from archaeology \cite{troiano2024comparative} and finance \cite{cai2016clustering} to industry \cite{lee2021technological} and zoology \cite{shen2021multivariate}, to name just a few.
-Especially in the field of unsupervised learning, clustering is utilized to extract structures from unlabeled data \cite{chander2023data}.
-
-Alongside the diverse applications of clustering, unique domain-specific requirements and challenges arise, which are addressed by leveraging different families of clustering approaches \cite{xu2015comprehensive,han2012data}.
-These approaches, ranging from partitioning and hierarchical methods to density-based, grid-based and graph-based methods, are designed to accommodate varying data characteristics and domain-related requirements to facilitate efficient cluster analyses across diverse applications.
-
-%different algorithms/problems
-%kmeans/lloyd
-%fast algorithms
-%inefficient: distance calculations
-%our proposal: ptolemaic
-%our purpose
-
-In this short paper, we focus on partitioning methods due to their simplicity in terms of interpretability and implementability \cite{DBLP:conf/iiwas/BeecksBHLSD22}.
-In this field, the \emph{k-means algorithm} \cite{bock2007clustering,hans2008origins,DBLP:journals/prl/Jain10,steinley2006k} has become one of the most influential clustering techniques \cite{DBLP:journals/kais/WuKQGYMMNLYZSHS08,olukanmi2019rethinking}.
-Although the term \emph{k-means algorithm} is explicitly credited to MacQueen \cite{macqueen1967}, its origins trace back to Steinhaus \cite{steinhaus1956division} in 1956, with its first application to data clustering by Forgy \cite{forgy1965cluster} in 1965.
-The widely recognized version in use today is the \emph{Lloyd algorithm} \cite{DBLP:journals/tit/Lloyd82}, introduced in 1982.
-
-With the proliferation of complex data sources and the growing demand for more efficient clustering techniques, numerous \emph{fast k-means algorithms} have been introduced \cite{DBLP:conf/icml/Elkan03,DBLP:conf/sdm/Hamerly10,drake2012accelerated,hamerly2015accelerating,DBLP:conf/icml/NewlingF16,DBLP:conf/icml/DingZSMM15,DBLP:conf/sisap/SchubertLF21,DBLP:conf/sisap/LangS23}.
-These methods are designed to achieve good clustering results with significantly reduced computational cost.
-Moreover, most of them do not require any form of precomputation, making them directly applicable across a wide range of scenarios.
-The major objective of fast k-means algorithms is to reduce the number of distance evaluations required when assigning data objects to cluster centers and to safely prune cluster centers from the assignment process.
-For this purpose, distances are approximated via lower and upper bounds.
-
-The Elkan algorithm \cite{DBLP:conf/icml/Elkan03} is a prominent representative of these algorithms and particularly suited for high-dimensional scenarios.
-It maintains one upper bound and multiple lower bounds for each data object in order to apply different pruning criteria for each combination of data object and cluster center.
-While the lower and upper bounds of Elkan's algorithm are directly derived from the triangle inequality, we propose to utilitze Ptolemy's inequality instead in order to increase the pruning performance.
-With this extension, we do not aim to compete with the state of the art in k-means clustering;
-rather, we intend to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations.
-We believe that our research findings offer a promising direction for future research and are beneficial for data scientists researchers and practitioners alike.
-
-This short paper is structured as follows:
-Section~\ref{Preliminaries} outlines the k-means problem and Elkan's approach of accelerating the algorithmic computation.
-In Section~\ref{sec:contrib}, we show how Ptolemy's inequality can be applied to the aforementioned algorithm. %as our main contribution.
-The results of our preliminary performance evaluation are detailed in Section~\ref{sec: results}, while Section~\ref{sec: conclusions} concludes this paper with a short outlook on future work.
-
-%contribution
-
-%structure
-
-%\newpage
-
-%\section{Related Work}
-%\todo{remove this section if we run out of time}
-
+\input{content/2_introduction}
 \input{content/3_preliminaries}
@@
 \input{content/6_conclusion.tex}
-\bibliography{references,bibliography_thesis,bibliography_iiwas}
+\bibliography{references/references,references/bibliography_thesis,references/bibliography_iiwas}
 \end{document}

### 2025-01-15 paper/content/2_introduction.tex
mention the original inspiration for the paper in the introduction

@@
 It maintains one upper bound and multiple lower bounds for each data object in order to apply different pruning criteria for each combination of data object and cluster center.
-While the lower and upper bounds of Elkan's algorithm are directly derived from the triangle inequality, we propose to utilitze Ptolemy's inequality instead in order to increase the pruning performance.
+Elkan's lower and upper bounds are directly derived from the triangle inequality; we propose to also utilize Ptolemy's inequality to increase the pruning performance, as this inequality has been successfully employed for database indexing \cite{hetlandPtolemaicAccessMethods2013,hetlandPtolemaicIndexing2015}.
 With this extension, we do not aim to compete with the state of the art in k-means clustering;

### 2025-02-27 paper/content/alg_elkan2.tex
fix inexact usage of the parameter in argmin

--- 2024-10-02 content/alg_elkan2.tex
@@
-\begin{algorithm}
+\begin{algorithm}[h]
 	\caption{Elkan's Algorithm}
@@
 For each point $x_i$ and each center $c_j$, set the lower bound $l(x_i,c_j)$ and the upper bound $u(x_i)$.
-Assign each $x_i$ to the nearest cluster $C_j$ such that $c(x_i) = \argmin_j d(x_i, c_j)$, utilizing Lemma 1 to minimize distance calculations.
-Set $r(x_i) = \text{true}$ for all points.
-		\todo{include Lemma 1}
+Assign each $x_i$ to the nearest cluster $C_j$ such that $c(x_i) = \argmin_{j'} d(x_i, c_{j'})$, utilizing that
+		$ d(c_j, c_m) \geq 2d(x_i, c_j) \Rightarrow d(x_i, c_m) \geq d(x_i, c_j) $
+		to minimize distance calculations. Set $r(x_i) = \text{true}$ for all points.
 
